---
title: 'UCI Online Shoppers Purchasing Intention Analysis'
author: hbmaclean
date: "`r Sys.Date()`"
editor_options:
  markdown:
    toc: true
    wrap: 72
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
---

```{css, echo = FALSE}
h1 {
    margin-top: 2em;
    border-bottom: 2px solid black;
    padding-bottom: 5px;
}
h2 {
    margin-top: 1em;
}
h3 {
    margin-top: 1em;
}
p {
    margin-top: 2px;
}
table {
    margin-top: 1em;
    margin-bottom: 1em;
    background: snow;
}
th {
    background: snow;
}
img {
    border: 1px solid black;
    margin-top: 4px;
    margin-bottom: 4px;
    width: 100%;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

I chose this dataset having worked in eCommerce. Below is a description
of the data from
[UCI](https://archive.ics.uci.edu/dataset/468/online+shoppers+purchasing+intention+dataset):

> The dataset consists of feature vectors belonging to 12,330 sessions.
> The dataset was formed so that each session would belong to a
> different user in a 1-year period to avoid any tendency to a specific
> campaign, special day, user profile, or period.
>
> The dataset consists of 10 numerical and 8 categorical attributes. The
> 'Revenue' attribute can be used as the class label.
>
> "Administrative", "Administrative Duration", "Informational",
> "Informational Duration", "Product Related" and "Product Related
> Duration" represent the number of different types of pages visited by
> the visitor in that session and total time spent in each of these page
> categories. The values of these features are derived from the URL
> information of the pages visited by the user and updated in real time
> when a user takes an action, e.g. moving from one page to another.
>
> The "Bounce Rate", "Exit Rate" and "Page Value" features represent the
> metrics measured by "Google Analytics" for each page in the e-commerce
> site. The value of "Bounce Rate" feature for a web page refers to the
> percentage of visitors who enter the site from that page and then
> leave ("bounce") without triggering any other requests to the
> analytics server during that session. The value of "Exit Rate" feature
> for a specific web page is calculated as for all pageviews to the
> page, the percentage that were the last in the session. The "Page
> Value" feature represents the average value for a web page that a user
> visited before completing an e-commerce transaction.
>
> The "Special Day" feature indicates the closeness of the site visiting
> time to a specific special day (e.g. Mother’s Day, Valentine's Day) in
> which the sessions are more likely to be finalized with transaction.
> The value of this attribute is determined by considering the dynamics
> of e-commerce such as the duration between the order date and delivery
> date. For example, for Valentina’s day, this value takes a nonzero
> value between February 2 and February 12, zero before and after this
> date unless it is close to another special day, and its maximum value
> of 1 on February 8.
>
> The dataset also includes operating system, browser, region, traffic
> type, visitor type as returning or new visitor, a Boolean value
> indicating whether the date of the visit is weekend, and month of the
> year.

# Installing Packages

```{r message = FALSE}
library(data.frame.X)
install_packages(c(
    'caret', 'stringr', 'readr', 'dplyr', 'kableExtra', 'DT', 'gridExtra',
    'ranger', 'randomForest', 'naivebayes', 'nnet', 'tibble', 'missMethods',
    'varhandle', 'devtools'
))

install_github('munoztd0/reprtree')
eval(substitute(library(pkg), list(pkg='reprtree')))
library(reprtree)

set.seed(234)
```

# Data Creation

This downloads the
[zip](https://archive.ics.uci.edu/dataset/73/mushroom), extracts the
data file, reads and parses the lines into a `data.frame` and adds
column names.

```{r message = FALSE}
# Download online+shoppers+purchasing+intention+dataset.zip from UCI.edu
# and render it as a data.frame
fetch_data <- function(verbose = FALSE) {
    options(timeout = 120)
    tmp   <- tempfile()
    zip   <- paste0(
        'https://archive.ics.uci.edu',
        '/static/public/468/online+shoppers+purchasing+intention+dataset.zip'
    )
    download.file(zip, tmp, quiet = !verbose)
    data <- read_csv(unz(tmp, 'online_shoppers_intention.csv')) 

    # Here I abbreviate some of the lengthy names as given
    colnames(data) <- c(
        'Admin', 'AdminDur', 'Info', 'InfoDur', 'Prod', 'ProdDur',
        'BounceRate', 'ExitRate', 'PageVal', 'SpecialDay', 'Month',
        'OS', 'Browser', 'Region', 'TrafficType', 'VisitorType',
        'Weekend', 'Revenue'
    )
    return(data)
}

shoppers_df <- fetch_data()
shoppers_df |>
    slice_sample(n = 10) |>
    mutate(across(where(is.numeric), ~ round(.x, 1))) |>
    datatable(
        rownames = FALSE,
        options  = list(
            dom        = 't',
            scrollX    = TRUE,
            pageLength = nrow(.data)
        )
    )
```

## Making Numeric

Making the few non-numeric columns numeric which will simplify analysis.
Month, obviously, can be converted to 1:12. The two logical columns,
Weekend and Revenue, can be represented as 1:0. VisitorType can be
represented by 1:3 for its three distinct values:

```{r}
table(shoppers_df$VisitorType) |>
    kable(col.names = c('VisitorType', 'Frequency')) |>
    kable_styling(full_width = FALSE, position = 'left')
```

```{r}
shoppers_df <- shoppers_df |>
    mutate(
        Month       = coalesce(match(Month, month.abb), match(Month,month.name)),
        VisitorType = match(VisitorType, as.vector(unique(VisitorType))),
        across(where(is.logical), ~ as.integer(.x))
    )
```

## Partitioning

```{r}
index     <- createDataPartition(y = shoppers_df$Revenue, times = 1, p = 0.5, list = FALSE)
train_df  <- shoppers_df[index,]
test_df   <- shoppers_df[-index,]
```

# Initial Analysis

## Class Distribution

85%/15% - may need to pick a model that supports class weights.

```{r}
round(prop.table(table(shoppers_df$Revenue)),2) |>
    kable(col.names = c('Revenue', 'Frequency')) |>
    kable_styling(full_width = FALSE, position = 'left')
```

## Missing, Duplicate, and Outlying Values {#missing_duplicate_outlying}

There is a small portion of duplicate rows. In my last project, I found
that removing duplicates crushed accuracy - so I will look to remove
these experimentally later.

```{r}
any_NA_dup_outliers(shoppers_df, tabular = TRUE)
```

There are *many* outliers and extreme values. I suspect most are
expected due to the nature of the data.

```{r}
colSums(rstatix::is_outlier(shoppers_df |> as.matrix())) |>
    as.data.frame() |>
    tibble::rownames_to_column(var = 'Column') |>
    rename(Outliers = c(2)) |>
    arrange(desc(Outliers)) |>
    inner_join(
        round(colMeans(shoppers_df),2) |>
        as.data.frame() |>
        tibble::rownames_to_column(var = 'Column') |>
        rename(Mean = c(2)),
        by = 'Column'
    ) |>
    inner_join(
        round(apply(shoppers_df, 2, FUN = sd),2) |>
        as.data.frame() |>
        tibble::rownames_to_column(var = 'Column') |>
        rename(SD = c(2)),
        by = 'Column'
    ) |>
    kable() |>
    kable_styling(
        bootstrap_options = 'bordered',
        full_width        = FALSE,
        position          = 'left'
    )
```

From work experience, I also suspect that durations are susceptible to
"cart abandonment" - shoppers leaving a session in a forgotten browser
window/tab and eventually timing out.

Coupling the information above with the bar charts below, it is clear
that product pages generally get lengthier looks - but also have extreme
outliers (presumed abandonment); while admin- and info pages generally
get brief looks - but also have outliers.

Note that the axes are scaled for better visibility of the thin end; and
that the Product Page axes have a 10-fold relationship with the other
durations. It does appear that the most extreme durations are associated
with zero revenue - consistent with a presumed session timeout.

```{r warning = FALSE}
common <- list(
    geom_bar(width = 5), scale_y_sqrt(), scale_x_sqrt(),
    labs(fill = 'Revenue', x = element_blank(), y = element_blank())
)
grid.arrange(
    shoppers_df |> ggplot(aes(ProdDur, fill = factor(Revenue))) +
        ggtitle('Product Page Durations') + common,
    shoppers_df |> ggplot(aes(AdminDur, fill = factor(Revenue))) +
        ggtitle('Admin Page Duration') + common,
    shoppers_df |> ggplot(aes(InfoDur, fill = factor(Revenue))) +
        ggtitle('Info Page Duration') + common,
    nrow = 3
)
```

Here is a second approach providing a similar look. Evidently, revenue
is best with very brief cumulative page views; but continues for product
page views up to about 15000 (ms, presumably.)

```{r}
grid.arrange(
    aggregate(shoppers_df, Revenue ~ ProdDur, sum) |>
        ggplot(aes(ProdDur, Revenue)) +
        geom_line(),
    aggregate(shoppers_df, Revenue ~ AdminDur, sum) |>
        ggplot(aes(AdminDur, Revenue)) +
        geom_line(),
    aggregate(shoppers_df, Revenue ~ InfoDur, sum) |>
        ggplot(aes(InfoDur, Revenue)) +
        geom_line(),
    nrow = 3
)
```

## Near-Zero Variance

As expected, durations and rates are variable; while other columns have
relatively few distinct values. None of the columns is considered to
have near-zero variance, and thus `nzv` and `zv` are not worthwhile
`preProcess` methods here.

```{r}
nzv_plot(shoppers_df)
```

Related to `nearZeroVar`, `checkConditonalX` reports any predictor that
is sparse conditioned on the levels of the outcome. Here, there is no
such predictor, and thus `conditionalX` is not a worthwhile `preProcess`
method here.

```{r}
checkConditionalX(shoppers_df |> select(-c(Revenue)), shoppers_df$Revenue)
```

## Correlation {#correlation}

Two pairs, `ExitRate` and `BounceRate`, and `Prod` and `ProdDur`, have
*about* 90% correlation. (Note that correlation is rounded to one digit
for legibility.) Perhaps one from each pair can be omitted.

`Browser`, `OS`, `Region`, and `Weekend`, and `TrafficType` to a lesser
extent, have noticeably weak correlations and sparse representation due
to high p-values. Perhaps one or more can be omitted.

It is also notable that `PageVal` has the largest correlation with
`Revenue`.

```{r}
corr_plot(shoppers_df, tl.srt = 90, digits = 1)
```

## Principal Component Analysis

The fact that these plots are difficult to read indicates that the
variance is spread widely and fairly evenly. `pca` does not look like a
worthwhile `preProcess` method here.

```{r fig.height = 5}
pca_plot(shoppers_df, pct = .90)
```

But oh my! Later I learned that [scale](#scale_and_center) is disastrous
here - causes `ranger` to always pick the negative class; and awhile
after learning that, I realized that my `pca_plot` had `scale = TRUE`
for `prcomp`. I've updated that to be optional, and... one PC expains
99% of the overall variance!

```{r}
pca_plot(shoppers_df, scale = FALSE, pct = .99)
```

Knowing that, I refer back to my coursework and its [Dimension
Reduction](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/dimension-reduction.html)
section, and plot the distance matrix for the data (a subset of it, for
performance reasons) versus the `prcomp` rotated data of the first PC.

```{r}
x        <- as.matrix(shoppers_df |> slice_sample(prop = .20))
pca      <- prcomp(x)
d        <- dist(x)
d_approx <- dist(pca$x[, 1])
plot(d, d_approx)
abline(0, 1, col = 'red')
```

## Quick Accuracy Check

Below is a method to check accuracy via `ranger`.

```{r}
vanilla_classification <- function(.train = train_df, .test = test_df, ...) {
    fit  <- ranger(factor(Revenue) ~ ., .train, ...)
    pred <- predict(fit, .test)
    cols <- c('Prevalence', 'Sensitivity', 'Specificity', 'Precision', 'F1')
    .cm  <- confusionMatrix(pred$predictions, factor(.test$Revenue))
    .cm$byClass |>
        as.data.frame() |>
        rownames_to_column(var = 'Class') |>
        tidyr::pivot_wider(names_from = c(1), values_from = c(2)) |>
        select(all_of(cols)) |>
        mutate(
            Accuracy = .cm$overall[['Accuracy']],
            across(everything(), ~ sprintf('%.2f', .x * 100))
        ) |>
        kable(
            digits  = 3,
            caption = 'Ranger Classification Metrics'
        ) |>
        kable_styling(
            bootstrap_options = 'bordered',
            full_width = FALSE,
            position = 'left'
        )
}
vanilla_classification()
```

90% seems good - but is only 5% better than simply guessing the negative
outcome.

## Positive Class {#positive_class}

By default, `TRUE`/`FALSE` and `1`/`0` are factored into the reverse
order; but in this case, logically, `Revenue=TRUE` is positive. Does the
order have any bearing?

```{r}
vanilla_classification(
    .train = train_df |> mutate(Revenue = reorder(factor(Revenue), Revenue, decreasing=TRUE)),
    .test  = test_df  |> mutate(Revenue = reorder(factor(Revenue), Revenue, decreasing=TRUE))
)
```

I think this is clarifying, so I will lock in that change.

-   Prevalence now reflects the positive class
-   Sensitivity and Specificity are in keeping with their definitions -
    True Positive and True Negative
-   F1 is much lower because it is based upon Sensitivity and Precision
-   Accuracy is essentially unchanged

```{r}
train_df <- train_df |>
    mutate(Revenue = reorder(factor(Revenue), Revenue, decreasing=TRUE))
test_df  <- test_df |>
    mutate(Revenue = reorder(factor(Revenue), Revenue, decreasing=TRUE))
```

## Omitting Columns {#omitting_columns}

Noted above in [Correlation](#correlation), `OS`, `Browser`, `Region`,
`Weekend`, and `TrafficType` have relatively few and weak correlations;
while `ExitRate` correlates strongly with `BounceRate`. I omitted
numerous combinations and find that these six can be dropped without
ill-effect - even perhaps a slight improvement, especially for
Specificity.

```{r}
train_df <- train_df |> select(-c(OS,Browser,Region,Weekend,TrafficType,ExitRate))
test_df  <- test_df  |> select(-c(OS,Browser,Region,Weekend,TrafficType,ExitRate))

vanilla_classification()
```

## Omitting Duplicate Rows {#omit_duplicates}

Noted above in [Missing, Duplicate, and Outlying
Values](#missing_duplicate_outlying), there are 125 duplicate rows. I
have learned to be cautious about removing them - and here again it
reduces accuracy.

```{r}
vanilla_classification(.train = distinct(train_df), .test = distinct(test_df))
```

## Class Weights {#class_weights}

For earlier projects, I've referred to [this
article](https://medium.com/@ravi.abhinav4/improving-class-imbalance-with-class-weights-in-machine-learning-af072fdd4aa4)
for weighing classes to overcome an imbalance. This does seem to help.
Numbers change with every run, but Sensitivity is generally improved -
which is precisely the goal, to more accurately predict true positives
despite the low prevalence.

```{r}
class_weights <- c(
    nrow(train_df) / (2 * sum(train_df$Revenue==1)),
    nrow(train_df) / (2 * sum(train_df$Revenue==0))
)
vanilla_classification(class.weights = class_weights)
```

## Case Weights {#case_weights}

There are also `case.weights`, for weighing the observations. I made two
educated guesses as to weighing the rows. The first method is
ineffective; the second method may be very slightly effective - in
reference to the results from switching the [Positive
Class](#positive_class).

1.  Sum across rows the values of three booleans - whether the
    Prod/Admin/Info page durations are outliers; then exaggerate the
    results from 0:3 to 1:20 with `exp`; and finally inverse the
    results, such that having no outlier equals 1 and having three
    outliers equals 1/20.

    ```{r}
    case_weights <- 1 / exp(
        rowSums(
            rstatix::is_outlier(
                train_df |>
                    select(c(ProdDur, AdminDur, InfoDur)) |>
                    as.matrix())
    ))
    vanilla_classification(case.weights = case_weights)
    ```

2.  Use `ntile` to spread the cumulative page durations into 50 buckets.

    ```{r}
    case_weights <- train_df |>
        mutate(DurationBucket = ntile(ProdDur+AdminDur+InfoDur, 50)) |> pull(DurationBucket)
    vanilla_classification(case.weights = case_weights)
    ```

## Impute Outliers {#impute_outliers}

Here I impute outliers with median values. It proves to be ineffective;
and factors are a mixed blessing...

```{r}
train_o      <- train_df |> mutate(Revenue = varhandle::unfactor(Revenue)) |> as.matrix()
outliers     <- rstatix::is_outlier(train_o[,c('ProdDur', 'InfoDur', 'AdminDur')])
idx          <- which(outliers)
train_o[idx] <- NA
train_o      <- train_o |>
    as.data.frame() |>
    mutate(Revenue = reorder(factor(Revenue), Revenue, decreasing=TRUE))

test_o       <- test_df |> mutate(Revenue = varhandle::unfactor(Revenue)) |> as.matrix()
outliers     <- rstatix::is_outlier(test_o[,c('ProdDur', 'InfoDur', 'AdminDur')])
idx          <- which(outliers)
test_o[idx]  <- NA
test_o       <- test_o |>
    as.data.frame() |>
    mutate(Revenue = reorder(factor(Revenue), Revenue, decreasing=TRUE))

vanilla_classification(
    .train = impute_median(train_o),
    .test  = impute_median(test_o)
)
```

## Scale and Center {#scale_and_center}

Heh, it seems scaling/centering causes `ranger` to always predict the
negative class. The same is true when just centering (`scale = FALSE`)
and just scaling (`center = FALSE`).

```{r}
vanilla_classification(
    .train = train_df |> mutate(across(where(is.numeric), ~ scale(.x)[1])),
    .test  = test_df  |> mutate(across(where(is.numeric), ~ scale(.x)[1]))
)
```

## Initial Conclusions {#initial_conclusions}

[Omitting columns](#omitting_columns) with weak correlations and high
p-values was NOT harmful, and thus considered worthwhile.

Switching the [positive class](#0) by reversing the natural factor order
of `Revenue` shifts and clarifies the metrics, particularly Prevalence,
Specificity and Sensitivity - without affecting accuracy.

[Case weights](#case_weights) are somewhat effective for balancing the
two outcomes.

[Omitting duplicates](#omit_duplicates), [imputing
outliers](#impute_outliers), and [scaling and
centering](#scale_and_center) are non-starters.

Looking ahead to `train` and the flexibility it provides, and probably
another model or few, I have already ruled out numerous `preProcess`
options:

-   `corr` (correlation)
-   `nzv`, `zv`, and `conditionalX` (variance-related)
-   `scale`, `center`
-   `median_impute` (NA'd outliers)

# Additional Models

Here I try some other classification models for comparison.

First, a configuration - for each model, a simple function/wrapper of
the model and the required `type` for `predict`.

```{r}
model_config <- list(
    ranger       = list(
        FUNC = function(...) { ranger(...) },
        TYPE = 'response'
    ),
    randomForest = list(
        FUNC = function(...) { randomForest(...) },
        TYPE = 'response'
    ),
    knn3         = list(
        FUNC = function(...) { knn3(...) },
        TYPE = 'class'
    ),
    naive_bayes  = list(
        FUNC = function(...) { naive_bayes(...) },
        TYPE = 'class'
    ),
    multinom     = list(
        FUNC = function(...) { multinom(...) },
        TYPE = 'class'
    )
)
```

And methods for predictions, wrapping `predict` with some intelligence
and error checking; and for accuracy, wrapping the call to `predict` and
`confusionMatrix` with error checking.

```{r}
predictions <- function(model, test_df, type = 'class') {
    tryCatch({
        .pred <- predict(model, test_df, type = type)
    }, 
    error = function(e) {
        print(c(Model = model, error = e))
    },
    finally = {
        if(!is.atomic(.pred)) {
            .pred <- .pred$predictions
        }
        return(.pred)
    })
}

accuracy <- function(model, test_df, ...) {
    tryCatch({
        .pred <- predictions(model, test_df, ...)
        .cm   <- confusionMatrix(.pred, as.factor(test_df$Revenue))
    },
    error = function(e) {
        print(c(Model = model, error = e))
    },
    finally = {
        if (!exists('.cm')) {
            return(NA)
        }
        else {
            return(round(100*.cm$overall[['Accuracy']],2))
        }
    })
}
```

And a "private" function to run a single model, with a caller to run all
models in the configuration.

```{r}
fc <- as.formula(factor(Revenue) ~ .)
.run_model <- function(cfg, .formula, .train, .test) {
    FUNC <- cfg$FUNC
    TYPE <- cfg$TYPE
    accuracy(
        get("FUNC")(.formula, .train),
        .test, type = TYPE
    )
}

run_models <- function(
        .models  = model_config,
        .formula = fc,
        .train   = train_df,
        .test    = test_df
    ) {
        results <- sapply(names(.models), function(model) {
            .run_model(.models[[model]], .formula, .train, .test)
        })
        names(results) <- names(.models)
        return(results)
}
```

Finally I run the models. It seems `ranger` was a fine first pick; with
`randomForest` keeping it company at about 90%.

```{r warning = FALSE}
shush <- capture.output({
    baselines <- run_models()
})
baselines |>
    as.data.frame() |>
    arrange(desc(baselines)) |>
    kable(col.names = c('Model', 'Accuracy')) |>
    kable_styling(full_width = FALSE, position = 'left')
```

# Training Several Models

Now I look to `train` and the flexibility it provides. Similar to
earlier, here is a "private" function to run a single model, and a
caller to run all models in the configuration.

```{r}
.train_model  <- function(
        model, cfg, .formula, .train, .test, bestTune,
       .trControl = NULL, .preProcess = NULL
) {
    tryCatch({
        fit <- (function(...) {
            caret::train(
                .formula,
                method     = model,
                data       = .train,
                trControl  = dplyr::coalesce(
                    .trControl,
                    trainControl(
                        method = 'repeatedcv',
                        number = 2,
                        p      = .5
                    )
                ),
                tuneGrid   = cfg$tuneGrid,
                preProcess = .preProcess,
                ...
            )
        })() # ideally, (...) 
    }, error = function(e) {
        print(c(Model = model, error = e))
    },
    finally = {
        if (exists('fit')) {
            .acc <- accuracy(fit, .test, type = dplyr::coalesce(cfg$type, 'raw'))

            if (bestTune) {
                return(
                    c(
                        Model    = model,
                        Accuracy = .acc,
                        bestTune = paste(
                            sapply(1:length(fit$bestTune), function(i) {
                                paste(
                                    names(fit$bestTune)[i],
                                    fit$bestTune[[i]],
                                    sep = '='
                                )
                            }),
                            collapse = ', '
                        )
                    )
                )
            }
            else {
                return(.acc)
            }
        }
    })
}

train_models <- function(
    .config       = train_config,
    .formula      = fc,
    .train        = train_df,
    .test         = test_df,
    bestTune      = FALSE,
    .trControl    = NULL,
    .preProcess   = NULL
) {
    results <- sapply(names(.config), function(model) {
        .train_model(
            model, .config[[model]], .formula, .train, .test,
            bestTune, .trControl, .preProcess
        )
    })
    names(results) <- names(.config)
    return(results)
}
```

## tuneGrid

And here, another configuration structure, with model-specific
`tuneGrid` objects - initially `NULL` for a baseline reference.

```{r}
train_config <- list(
    ranger = list(
        tuneGrid = NULL
    ),
    rf = list(
        tuneGrid = NULL
    ),
    Rborist = list(
        tuneGrid = NULL
    ),
    treebag = list(
        tuneGrid = NULL
    ),
    xgbTree = list(
        tuneGrid = NULL
    ),
    svmRadial = list(
        tuneGrid = NULL
    )
)
```

And finally, a baseline run:

```{r message = FALSE, warning = FALSE}
fc <- as.formula(factor(Revenue) ~ .)
shush <- capture.output({
    train_baselines <- train_models()
})
train_baselines |>
    as.data.frame() |>
    arrange(desc(train_baselines)) |>
    kable(col.names = c('Model', 'Accuracy')) |>
    kable_styling(full_width = FALSE, position = 'left')
```

Now another run to capture each model's `bestTune`:

```{r warning = FALSE}
shush <- capture.output({
    bestTunes <- train_models(bestTune = TRUE)
})

t(unname(bestTunes)) |>
    kable(col.names = c('Model', 'Accuracy', 'bestTune')) |>
    kable_styling(
        full_width = FALSE,
        position   = 'left',
        bootstrap_options = c('bordered', 'striped')
    )
```

Now another run with `tuneGrid` parameters spanning the initial best
values. Having used `xgbTree` before, I know to be careful with its
sprawling `tuneGrid` - to not create too many permutations.

```{r}
train_config <- list(
    ranger = list(
        tuneGrid = expand.grid(
            mtry          = c(2,4,6,8),
            splitrule     = c('gini', 'extratrees'),
            min.node.size = c(1,5,10,20)
        )
    ),
    rf = list(
        tuneGrid = expand.grid(
            mtry = c(2,4,6,8)
        )
    ),
    Rborist = list(
        tuneGrid = expand.grid(
            predFixed = c(2,4,6,8),
            minNode   = c(1,2,5,10)
        )
    ),
    treebag = list(
        tuneGrid = NULL
    ),
    xgbTree = list(
        tuneGrid = expand.grid(
            nrounds          = 50,
            max_depth        = c(2,4,10),
            eta              = c(0.01, 0.1, 0.4),
            gamma            = c(0, .3),
            colsample_bytree = 0.6,
            min_child_weight = c(1,3,5),
            subsample        = 1
        )
    ),
    svmRadial = list(
        tuneGrid = expand.grid(
            sigma = c(.001, .01, .1, .3),
            C     = c(.5, 1, 5, 10)
        )
    )
)
```

Well, I have never really had much luck making improvements via
`tuneGrid`. I suspect models already do the same - try *ranges* of
values and hone in on the best. Some values did change, but accuracy was
never improved significantly. Fruitless again, other than *maybe*
`xgbTree`, and figuring out a solution for presenting `bestTune` values.

```{r warnings = FALSE, message = FALSE}
shush <- capture.output({
    bestTunesII <- train_models(bestTune = TRUE)
})

t(unname(bestTunesII)) |>
    kable(col.names = c('Model', 'Accuracy', 'bestTune')) |>
    kable_styling(
        full_width = FALSE,
        position   = 'left',
        bootstrap_options = c('bordered', 'striped')
    )
```

I'll lock in the settings for `xgbTree` and `Rborist` and pitch the
rest.

```{r}
train_config <- list(
    ranger = list(
        tuneGrid = NULL
    ),
    rf = list(
        tuneGrid = NULL
    ),
    Rborist = list(
        tuneGrid = expand.grid(
            predFixed = 2,
            minNode   = c(10,20,30)
        )
    ),
    treebag = list(
        tuneGrid = NULL
    ),
    xgbTree = list(
        tuneGrid = expand.grid(
            nrounds          = 50,
            max_depth        = 4,
            eta              = 0.1,
            gamma            = c(.3,.5,.7),
            colsample_bytree = 0.6,
            min_child_weight = 3,
            subsample        = 1
        )
    ),
    svmRadial = list(
        tuneGrid = NULL
    )
)
```

```{r warnings = FALSE, message = FALSE}
shush <- capture.output({
    bestTunesIII <- train_models(bestTune = TRUE)
})

t(unname(bestTunesIII)) |>
    kable(col.names = c('Model', 'Accuracy', 'bestTune')) |>
    kable_styling(
        full_width = FALSE,
        position   = 'left',
        bootstrap_options = c('bordered', 'striped')
    )
```

## preProcess

I have already ruled out many `preProcess` options in my [initial
conclusions](#initial_conclusions). Here I try `BoxCox`, `YeoJohnson`,
`spacialSign` and `pca`. I tried `expoTrans` as well, but it spewed
*Expo. transform induced infinite values in several predictors and is
ommitted: AdminDur, InfoDur, ProdDur*.

```{r}
display_summary <- function(.data) {
    .data |>
        dplyr::rename(` ` = c(1)) |>
        datatable(
            rownames = FALSE,
            options  = list(
                dom        = 't',
                pageLength = nrow(.data)
            )
        )
}
```

```{r warning = FALSE, message = FALSE, error = FALSE}
shush <- capture.output({
    pre_results <- bind_rows(
        train_models(.preProcess = 'BoxCox'),
        train_models(.preProcess = 'YeoJohnson'),
        train_models(.preProcess = 'spatialSign'),
        train_models(.preProcess = 'pca',
            .trControl = trainControl(
                preProcOptions = list(thresh = .99),
                method = 'repeatedcv',
                number = 2,
                p      = .5
            )
        )
    )
})

pre_summary <- bind_cols(
    c('BoxCox', 'YeoJohnson', 'spatialSign', 'pca'),
    pre_results
)

display_summary(pre_summary)
```

## trainControl

```{r warning = FALSE, message = FALSE}
boot_results <- bind_rows(
    train_models(.trControl = trainControl(
        method = 'boot', number = 2)),
    train_models(.trControl = trainControl(
        method = 'boot632', number = 2)),
    train_models(.trControl = trainControl(
        method = 'optimism_boot', number = 2)),
    train_models(.trControl = trainControl(
        method = 'boot_all', number = 2))
)

cv_results <- bind_rows(
    train_models(.trControl = trainControl(
        method = 'cv', p = .5, repeats = 2, number = 2)),
    train_models(.trControl = trainControl(
        method = 'repeatedcv', p = .5, repeats = 2, number = 2)),
    train_models(.trControl = trainControl(
        method = 'LGOCV', p = .5, repeats = 2, number = 2))
)

train_summary <- bind_cols(
    c(
        'boot', 'boot632', 'optimism_boot', 'boot_all',
        'cv', 'repeatedcv', 'LGOCV'
    ),
    bind_rows(boot_results, cv_results)
)

display_summary(train_summary)
```

# Focus on xgbTree

I do not know how authoritative this
[visual-xgboost-tuning-with-caret](https://www.kaggle.com/code/pelkoja/visual-xgboost-tuning-with-caret)
article is, but it seems knowledgeable. Here, I follow it quite closely,
in my own style.

```{r warning = FALSE}
shush <- capture.output({
    xgb <- caret::train(
        fc,
        method     = 'xgbTree',
        data       = train_df,
        trControl  = trainControl(
                        method = 'cv',
                        number = 3
                    ),
        tuneGrid   = expand.grid(
            nrounds          = seq(200, 1000, 50),
            eta              = c(0.025, 0.05, 0.1, 0.3),
            max_depth        = 2:6,
            min_child_weight = 1,
            colsample_bytree = 1,
            subsample        = 1,
            gamma            = 0
        )
    )
})
```

Here I plot accuracy per `max_depth` versus `nrounds`, faceted by `eta`.

```{r}
xgb$results |>
    ggplot(aes(nrounds, Accuracy, color = factor(max_depth))) +
    geom_point() +
    geom_line() +
    facet_wrap(~eta) + 
    labs(color = 'Max Tree Depth') +
    ggtitle(
        'Accuracy vs. N-Rounds per Tree Depth, Faceted by eta',
        subtitle = sprintf('Best Accuracy, %.2f', max(xgb$results$Accuracy)*100)
    )
```

Best accuracy above should align with `bestTune`:

```{r}
xgb$bestTune |>
    as_tibble() |>
    kable(caption = sprintf('Accuracy %%%.2f', max(xgb$results$Accuracy)*100)) |>
    kable_styling(bootstrap_options = 'bordered')
```

Here I lock in the "best" `eta`; continue flexing on `max_depth`; and
add variability for `min_child_weight`.

```{r warning = FALSE}
eta   <- xgb$bestTune$eta
shush <- capture.output({
    xgb <- caret::train(
        fc,
        method     = 'xgbTree',
        data       = train_df,
        trControl  = trainControl(
                        method = 'cv',
                        number = 3
                    ),
        tuneGrid   = expand.grid(
            nrounds          = seq(200, 1000, 50),
            eta              = eta,
            max_depth        = 2:5,
            min_child_weight = 1:4,
            colsample_bytree = 1,
            subsample        = 1,
            gamma            = 0
        )
    )
})
```

Here, I plot accuracy per `min_child_weight` versus `nrounds`, faceted
by `max_depth`.

```{r}
xgb$results |>
    ggplot(aes(nrounds, Accuracy, color = factor(min_child_weight))) +
    geom_point() +
    geom_line() +
    facet_wrap(~max_depth) + 
    labs(color = 'Min Child Weight') +
    ggtitle(
        'Accuracy vs. N-Rounds per Child Weight, Faceted by Max Depth',
        subtitle = sprintf('Best Accuracy, %.2f', max(xgb$results$Accuracy)*100)
    )
```

And inspect `bestTune`:

```{r}
xgb$bestTune |>
    as_tibble() |>
    kable(caption = sprintf('Accuracy %%%.2f', max(xgb$results$Accuracy)*100)) |>
    kable_styling(bootstrap_options = 'bordered')
```

Here I lock in "best" `max_depth` and `min_child_weight`, along with
`eta`; and create sequences for `colsample_bytree` and `subsample`.
(Again, being cautious about the size of `tuneGrid` - here, 250
permutations.)

```{r}
max_depth        <- xgb$bestTune$max_depth
min_child_weight <- xgb$bestTune$min_child_weight

shush <- capture.output({
    xgb <- caret::train(
        fc,
        method     = 'xgbTree',
        data       = train_df,
        trControl  = trainControl(
                        method = 'cv',
                        number = 3
                    ),
        tuneGrid   = expand.grid(
            nrounds          = seq(200, 1000, 50),
            eta              = eta,
            max_depth        = max_depth,
            min_child_weight = min_child_weight,
            colsample_bytree = seq(.4, 1, .2),
            subsample        = seq(.2, 1, .2),
            gamma            = 0
        )
    )
})
```

And here, I plot accuracy per `subsample` versus `nrounds`, faceted by
`colsample_bytree`.

```{r}
xgb$results |>
    ggplot(aes(nrounds, Accuracy, color = factor(subsample))) +
    geom_point() +
    geom_line() +
    facet_wrap(~colsample_bytree) + 
    labs(color = 'Subsample') +
    ggtitle(
        'Accuracy vs. N-Rounds per Subsample, Faceted by colsample_bytree',
         subtitle = sprintf('Best Accuracy, %.2f', max(xgb$results$Accuracy)*100)
)
```

And inspect `bestTune`:

```{r}
xgb$bestTune |>
    as_tibble() |>
    kable(caption = sprintf('Accuracy %%%.2f', max(xgb$results$Accuracy)*100)) |>
    kable_styling(bootstrap_options = 'bordered')
```

Here I lock in the "best" `colsample_bytree` and `subsample`, and create
a sequence for `gamma`.

```{r}
colsample_bytree <- xgb$bestTune$colsample_bytree
subsample        <- xgb$bestTune$subsample

shush <- capture.output({
    xgb <- caret::train(
        fc,
        method     = 'xgbTree',
        data       = train_df,
        trControl  = trainControl(
                        method = 'cv',
                        number = 3
                    ),
        tuneGrid   = expand.grid(
            nrounds          = seq(200, 1000, 50),
            eta              = eta,
            max_depth        = max_depth,
            min_child_weight = min_child_weight,
            colsample_bytree = colsample_bytree,
            subsample        = subsample,
            gamma            = seq(0, 1, .1)
        )
    )
})
```

And plot accuracy per `gamma` versus `nrounds`.

```{r}
xgb$results |>
    ggplot(aes(nrounds, Accuracy, color = factor(gamma))) +
    geom_point() +
    geom_line() +
    labs(color = 'Gamma') +
    ggtitle(
        'Accuracy vs. N-Rounds per Gamma',
         subtitle = sprintf('Best Accuracy, %.2f', max(xgb$results$Accuracy)*100)
)
```

Inspect `bestTune`:

```{r}
xgb$bestTune |>
    as_tibble() |>
    kable(caption = sprintf('Accuracy %%%.2f', max(xgb$results$Accuracy)*100)) |>
    kable_styling(bootstrap_options = 'bordered')
```

Well, that does give the best accuracy during this process - this time,
but the results are ever-changing despite seed. Here I lock in the
"best" `gamma`, and incorporate 'YeoJohnson' as a `preProcess` method,
and replace 'cv' with 'optimism_boot' for the `trainControl` method -
two items that made minor improvements for this model earlier.

```{r}
gamma <- xgb$bestTune$gamma

shush <- capture.output({
    xgb <- caret::train(
        fc,
        method     = 'xgbTree',
        data       = train_df,
        trControl  = trainControl(method = 'optimism_boot'),
        tuneGrid   = expand.grid(
            nrounds          = seq(200, 1000, 50),
            eta              = eta,
            max_depth        = max_depth,
            min_child_weight = min_child_weight,
            colsample_bytree = colsample_bytree,
            subsample        = subsample,
            gamma            = gamma
        ),
        preProcess = c('YeoJohnson')
    )
})
```

Inspect `bestTune` and resulting accuracy:

```{r}
xgb$bestTune |>
    as_tibble() |>
    kable(caption = sprintf('Accuracy %%%.2f', max(xgb$results$Accuracy)*100)) |>
    kable_styling(bootstrap_options = 'bordered')
```

```{r}
nrounds <- xgb$bestTune$nrounds
eta     <- xgb$bestTune$eta

shush <- capture.output({
    xgb <- caret::train(
        fc,
        method     = 'xgbTree',
        data       = train_df,
        trControl  = trainControl(
            method         = 'optimism_boot',
            preProcOptions = list(thresh = .99)
        ),
        tuneGrid   = expand.grid(
            nrounds          = nrounds,
            eta              = eta,
            max_depth        = max_depth,
            min_child_weight = min_child_weight,
            colsample_bytree = colsample_bytree,
            subsample        = subsample,
            gamma            = gamma
        ),
        preProcess = c('YeoJohnson', 'pca')
    )
})

xgb$bestTune |>
    as_tibble() |>
    kable(caption = sprintf('Accuracy %%%.2f', max(xgb$results$Accuracy)*100)) |>
    kable_styling(bootstrap_options = 'bordered')
```

# Decision Tree

I like to finish with a glimpse at a decision tree, for a sense of the
inner mechanics.

```{r fig.height = 14, fig.width = 14}
fit                   <- randomForest(Revenue ~ ., train_df, ntree = 20)
simplest_tree         <- which.min(fit$forest$ndbigtree)
fit$forest$xbestsplit <- round(fit$forest$xbestsplit, 2)  # 'digits' workaround
plot.getTree(
    fit,
    k     = simplest_tree,
    depth = 6,
    main  = sprintf(
         'RandomForest Decision Tree - Accuracy %.2f',
         accuracy(fit, test_df)
    )
)
```

# Conclusions

Again, my primary goal was continued learning. I used `t` to rotate
`bestTune` objects, after pondering `pivot_wider` and `pivot_longer`;
discovered `capture.output` to silence unwanted messages; had my first
opportunity since classwork to look at `dist` (for other projects, it
ran out of memory or simply seemed inapplicable); and realized my error
of assuming `scale=TRUE` for `prcomp`.

Once again though, while I am able to fairly quickly identify one good
model from many, I seem unable to make measurable improvements with
`preProcess`, `trainControl`, `tuneGrid`, imputing outliers, reducing
dimensions, or other attempts.

Four models slightly exceeded 90%, but none consistently stood out.

In this case, I *pause* with a working model; disappointed that it is
merely 5% better than guessing, and frustrated that I could not overcome
the class imbalance.

# References

<https://www.kaggle.com/code/pelkoja/visual-xgboost-tuning-with-caret>

<https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/dimension-reduction.html>
