---
title: "Kaggle Twitter Sentiment Analysis"
author: "hbmaclean"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
editor_options:
  markdown:
    toc: true
    wrap: 72
---

```{css, echo = FALSE}
h1 {
    margin-top: 2em;
    border-bottom: 2px solid black;
    padding-bottom: 5px;
}
h2 {
    margin-top: 1em;
}
h3 {
    margin-top: 1em;
}
p {
    margin-top: 2px;
}
table {
    margin-top: 1em;
    margin-bottom: 1em;
    background: snow;
}
th {
    background: snow;
}
img {
    border: 1px solid black;
    margin-top: 4px;
    margin-bottom: 4px;
    width: 100%;
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Below is a description of the **Twitter Sentiment Analysis Dataset**
from
[Kaggle](https://www.kaggle.com/datasets/jp797498e/twitter-entity-sentiment-analysis/data).

> **Overview**
>
> This is an entity-level sentiment analysis dataset of twitter. Given a
> message and an entity, the task is to judge the sentiment of the
> message about the entity. There are three classes in this dataset:
> Positive, Negative and Neutral. We regard messages that are not
> relevant to the entity (i.e. Irrelevant) as Neutral.
>
> **Usage**
>
> Please use `twitter_training.csv` as the training set
> and `twitter_validation.csv` as the validation set. Top 1
> classification accuracy is used as the metric.

# Importing Packages

```{r warning = FALSE, message = FALSE}
library(data.frame.X)
install_packages(c(
    'curl', 'stringr', 'dplyr', 'tidyr', 'caret', 'tibble', 'kableExtra', 'DT',
    'syuzhet', 'ranger', 'randomForest', 'devtools'
))

install_github('munoztd0/reprtree')
eval(substitute(library(pkg), list(pkg='reprtree')))
library(reprtree)

set.seed(45245)
```

# Fetching the Data

In this case, the data is pre-partitioned into training and validation.
This downloads the zip and extracts and reads each CSV into a
data.frame. Given the note in the data description above, this also
mutates "irrelevant" to "neutral".

```{r}
fetch_data <- function() {
    options(timeout = 120)
    url <- paste0(
        'https://www.kaggle.com/api/v1/datasets/download/',
        'jp797498e/twitter-entity-sentiment-analysis'
    )
    tmp <- tempfile()
    curl_download(url, tmp)
    cols <- c('id', 'entity', 'sentiment', 'content')
    train_df <<- as.data.frame(
        read.csv(
            unz(tmp, 'twitter_training.csv'),
            header = FALSE,
            col.names = cols
        )
    ) |>
        mutate(sentiment = ifelse(sentiment == 'Irrelevant', 'Neutral', sentiment))

    test_df <<- as.data.frame(
        read.csv(
            unz(tmp, 'twitter_validation.csv'),
            header = FALSE,
            col.names = cols
        )
    ) |>
        mutate(sentiment = ifelse(sentiment == 'Irrelevant', 'Neutral', sentiment))

}
fetch_data()
sample_n(test_df, 5) |> union_all(sample_n(train_df, 5)) |>
    datatable(rownames = FALSE, options = list(dom = 't', pageLength = nrow(.data)))
```

# Initial Data Analysis

Here I run through the typical first inspection.

## Check for NAs, Duplicates, and Outliers

Are there any NAs, duplicates, outliers, or extreme values?

```{r warning = FALSE}
any_NA_dup_outliers(train_df, tabular = TRUE)
```

Ah hold on, the `id` column surely has no predictive value, and has such
a range that it will inherently have outliers and extreme values, and
perhaps duplicates. I will remove that column and try again:

```{r warning = FALSE}
test_df$id  <- NULL
train_df$id <- NULL
any_NA_dup_outliers(train_df) |>
    add_row(any_NA_dup_outliers((test_df))) |>
    kable(align = 'c') |>
    kable_styling(
        bootstrap_options = c('bordered'),
        full_width = FALSE, position = 'left'
    ) |>
    pack_rows(
        index = c('train' = 1, 'test' = 1),
        label_row_css = 'text-align:center; background-color: #666; color: #fff;'
    )
```

The training set has duplicates to be purged.

```{r}
train_df  <- distinct(train_df)
```

## PCA, Correlation, and Variation

With the three remaining columns, it is clear that the tweet content
bears all significant variability and predictive power - I will not even
look at PCA, correlation, or variation.

However, I assume that one tweet word may have different sentiment
values for different entities - e.g. that "murder" is evidently positive
for Borderlands, but presumably negative for Amazon, etc. I am assuming
a correlation between `entity` and the sentiments of `content`, but at a
word-level that I cannot simply check with `corr`. I will call this the
entity-effect.

## Simple Classification Check

Here I arbitrarily pick `ranger` as a classification model and run it as
plain as can be (as a function, to be called again later.)

```{r}
vanilla_classification <- function(.train = train_df, .test = test_df) {
    etime <- system.time(
        fit  <- ranger(factor(sentiment) ~ ., .train)
    )
    pred <- predict(fit, .test)
    
    cols <- c('Class', 'Prevalence', 'Sensitivity', 'Specificity', 'Precision', 'F1')
    .cm  <- confusionMatrix(pred$predictions, factor(.test$sentiment))
    .cm$byClass |>
        as.data.frame() |>
        rownames_to_column(var = 'Class') |>
        select(all_of(cols)) |>
        kable(
            digits  = 3,
            caption = sprintf(
                'Ranger Classification Metrics - Accuracy %%%.2f; ranger time %.2f(s)',
                .cm$overall[['Accuracy']]*100,
                etime['elapsed']
            )
        ) |>
        kable_styling(
            bootstrap_options = 'bordered',
            full_width = FALSE,
            position = 'left'
        )
}
vanilla_classification()
```

Almost 60% - almost twice the chance of guessing; with an interesting
imbalance between sensitivity and specificity.

# Tweet Content - Lexicon Sentiments

The tweet content looks pretty messy, especially URLs (without protocol,
and often broken up by white space, e.g. "dlvr.it / RMTrgF". Before
attempting to clean up with regular expressions, I'll see how I fare
with the data as-is.

In my coursework, I learned of `tidytext`, and its `unnest_tokens`,
`stop_words` and `get_sentiments` to extract tokens/words from a
string/column, omit those of little value, and get a "sentiment" for
each via one of several lexicons. That was a fine way of teaching what
is involved, but in reviewing for this task I found [Sentiment Analysis
(Lexicons)](https://rpubs.com/chelseyhill/676279) and a means for
getting one sentiment for an entire block of text.

I try that below - getting one sentiment per tweet, from four lexicons.
I also calculate a normalized consensus sentiment, deviating slightly
from the approach in the article.

```{r}
train_df <- train_df |>
    mutate(
        bing      = syuzhet::get_sentiment(content, method = 'bing'),
        afinn     = syuzhet::get_sentiment(content, method = 'afinn'),
        nrc       = syuzhet::get_sentiment(content, method = 'nrc'),
        syuzhet   = syuzhet::get_sentiment(content, method = 'syuzhet'),
        consensus = sign(
            sign(bing) + sign(afinn) + sign(nrc) + sign(syuzhet)
        )
    )

test_df <- test_df |>
    mutate(
        bing      = syuzhet::get_sentiment(content, method = 'bing'),
        afinn     = syuzhet::get_sentiment(content, method = 'afinn'),
        nrc       = syuzhet::get_sentiment(content, method = 'nrc'),
        syuzhet   = syuzhet::get_sentiment(content, method = 'syuzhet'),
        consensus = sign(
            sign(bing) + sign(afinn) + sign(nrc) + sign(syuzhet)
        )
    )
```

And now to see what difference that made:

```{r}
vanilla_classification()
```

Much improved accuracy; much better balanced sensitivity and
specificity; and/but longer runtime, as expected.

Note:

-   `syuzhet` also supports a `stanford` lexicon, for which you must
    install `coreNLP` and provide its path via 'path_to_tagger'.
-   Evidently "syuzhet" is a Russian term for "plot" or "storyline".

## PCA and Correlation!

With the additional columns, PCA (`prcomp`) and Correlation (`cor`) get
more interesting. Granted, there are not many columns, but two PCs
account for 85% of the overall variance.

```{r warning = FALSE}
pca_plot(train_df, pct = .85, which = 'bar')
```

Two columns, `content` and `entity`, have a 97% correlation. I had
already hoped that `content` is unnecessary with the additional
sentiment columns. `bing`, `syuzhet`, and `afinn` all correlate strongly
(80+%), `nrc` appears to be an outsider. Alas, evidently nothing
correlates strongly with `sentiment`.

```{r warning = FALSE}
corr_plot(train_df)
```

Conclusion: `pca` and `corr` may be good `preProcess` options for
`train`. Defaults for `preProcess` are `thresh = .95` for `pca` and
`cutoff = .9` for `corr`. I will also consider dropping `nrc`, or at
least omitting it via formula; and the same with `content`.

# Data Cleanup

Here I convert the remaining non-numeric data to numbers; check whether
`content` is still necessary given the summarized lexicon sentiments;
whether all the new lexicon sentiment columns are necessary; and check
again for resulting duplicates.

## Making it Numeric

I have cycled through this process numerous times. Once, dropping
`content` had comparable accuracy, so I did. Another time, it lowered
accuracy by %10 so I did not - but then hit the 16GB vector limit. Here
I drop it regardless, and then convert `entitity` and `sentiment` to
numeric.

While at it, I round `syuzhet` to three decimals, which in turn
introduces tens of duplicates.

```{r}
unq_sentiments <- unique(c(train_df$sentiment, test_df$sentiment))
unq_entities   <- unique(c(train_df$entity, test_df$entity))

train_df <- train_df |> mutate(
    content   = NULL,
    sentiment = match(sentiment, unq_sentiments) -2,
    entity    = match(entity, unq_entities),
    syuzhet   = round(syuzhet, 3)
)
test_df <-test_df |> mutate(
    content   = NULL,
    sentiment = match(sentiment, unq_sentiments) -2,
    entity    = match(entity, unq_entities),
    syuzhet   = round(syuzhet, 3)
)

vanilla_classification()
```

## Partitioning

Here I partition into train/test, after tucking aside full copies for
later. Note, it is extremely tempting to drop duplicate rows, now about
53%(!) but accuracy plummets.

```{r}
train_org <- train_df
test_org  <- test_df

index     <- createDataPartition(y = train_df$sentiment, times = 1, p = 0.5, list = FALSE)
train_df  <- train_df[index,]
test_df   <- train_df[-index,]

vanilla_classification()
```

Faster, obviously; and more accurate (against partitioned half vs.
actual test set.)

## Accuracy without 'nrc'

```{r}
vanilla_classification(
    .train = train_df |> select(-c(nrc)),
    .test =  test_df  |> select(-c(nrc))
)
```

## Accuracy without 'afinn'

```{r}
vanilla_classification(
    .train = train_df |> select(-c(afinn)),
    .test =  test_df  |> select(-c(afinn)))
```

## Accuracy without 'bing'

```{r}
vanilla_classification(
    .train = train_df |> select(-c(bing)),
    .test =  test_df  |> select(-c(bing))
)
```

## Accuracy without 'syuzhet'

```{r}
vanilla_classification(
    .train = train_df |> select(-c(syuzhet)),
    .test =  test_df  |> select(-c(syuzhet))
)
```

## Accuracy without 'consensus'

```{r}
vanilla_classification(
    .train = train_df |> select(-c(consensus)),
    .test  = test_df  |> select(-c(consensus))
)
```

## Dropping 'bing' and 'consensus'

Accuracy is stable to improved without `bing` and `consensus` - so herte
I drop them.

```{r}
train_df$bing  <- train_df$consensus  <- NULL
test_df$bing   <- test_df$consensus   <- NULL
train_org$bing <- train_org$consensus <- NULL
test_org$bing  <- test_org$consensus  <- NULL

vanilla_classification()
```

# Initial Multi-Model Framework

Here I create functions for making predictions, calculating accuracy,
running individual models, tracking results, and easily running them all
together. The goal is to run numerous models and then narrow down to the
most accurate one or few.

## Predictions and Accuracy

Below are functions to make predictions and calculate accuracy, using
the
[ellipsis](https://cran.r-project.org/web/packages/ellipsis/ellipsis.pdf)
argument to allow model nuances to be passed in.

```{r}
predictions <- function(model, test_df, type = 'class') {
    tryCatch({
        .pred <- predict(model, test_df, type = type)
    }, 
    error = function(e) {
        print(c(Model = model, error = e))
    },
    finally = {
        if(!is.atomic(.pred)) {
            .pred <- .pred$predictions
        }
        return(.pred)
    })
}

accuracy <- function(model, test_df, ...) {
    tryCatch({
        .pred <- predictions(model, test_df, ...)
        .cm   <- confusionMatrix(.pred, as.factor(test_df$sentiment))
    },
    error = function(e) {
        print(c(Model = model, error = e))
    },
    finally = {
        if (!exists('.cm')) {
            return(NA)
        }
        else {
            return(round(100*.cm$overall[['Accuracy']],2))
        }
    })
}
```

## Formula and Models

And here, a basic formula; and a function for each model plus the type
for the underlying `predict` - generally 'class' or 'response'.

```{r}
fc <- as.formula(factor(sentiment) ~ .)
models <- list(
    ranger       = list(
        FUNC = function(...) { ranger(...) },
        TYPE = 'response'
    ),
    randomForest = list(
        FUNC = function(...) { randomForest(..., ntree = 20) },
        TYPE = 'response'
    ),
    knn3         = list(
        FUNC = function(...) { knn3(...) },
        TYPE = 'class'
    )
)
```

**Notes:**

1.  Originally, I had six models - the above three plus `multinom`,
    `naive_bayes`, and `avNNet`. I eventually pared it down to shave
    execution time.
2.  `ranger` was uniquely problematic here. Originally, rather than
    `TYPE`, I set this up as `extraArgs` with a value of
    `c(type = 'class')` for example, or `NULL` if the underlying default
    was sufficient. But `ranger` gave, *Error: Expecting a single value:
    [extent=0]* when `extraArgs` was `NULL`; and *Error: Not compatible
    with requested type: [type=character; target=logical]* when
    `extraArgs` was `c(type = 'response')`. All other models behaved as
    expected, with a key-value pair or `NULL`.

## Model Runner

Finally, a function to actually run the models and accumulate metrics.

```{r}
.run_model <- function(cfg, .formula, .train, .test) {
    FUNC <- cfg$FUNC
    TYPE <- cfg$TYPE
    accuracy(
        get("FUNC")(.formula, .train),
        .test, type = TYPE
    )
}

run_models <- function(
        .models  = models,
        .formula = fc,
        .train   = train_df,
        .test    = test_df
    ) {
        results <- sapply(names(.models), function(model) {
            .run_model(.models[[model]], .formula, .train, .test)
        })
        names(results) <- names(.models)
        return(results)
}
```

## Metrics Summary

Below, a function to display the metrics. At the moment, the expectation
is that the named vector from `run_models` will be bound with an initial
column of meaningful labels.

```{r}
display_summary <- function(.data) {
    .data |>
        dplyr::rename(` ` = c(1)) |>
        datatable(
            rownames = FALSE,
            options  = list(
                dom        = 't',
                pageLength = nrow(.data)
            )
        )
}
```

## Initial Runs - Baseline, Scaling, Centering

With the framework defined, I run the models again with the lexicon
columns scaled, centered, and scaled+centered.

```{r warning = FALSE, message = FALSE}
baselines <- run_models()
baselines |>
    as.data.frame() |>
    arrange(desc(baselines)) |>
    kable(col.names = c('Model', 'Accuracy')) |>
    kable_styling(full_width = FALSE, position = 'left')
```

# Second Multi-Model Framework - train()

## Models and Tuning

I will stick with `ranger` and add in `Rborist` and `treebag`.

First, I define a list of NULL `trControl`, `tuneGrid`, and `preProcess`
objects (and maybe `extraArgs`, but there is a rant coming up.)

```{r}
train_config <- list(
    ranger = list(
        trControl  = NULL,
        tuneGrid   = NULL,
        preProcess = NULL,
        extraArgs  = NULL
    ),
    Rborist = list(
        trControl  = NULL,
        tuneGrid   = NULL,
        preProcess = NULL,
        extraArgs  = NULL
    ),
    treebag = list(
        trControl  = NULL,
        tuneGrid   = NULL,
        preProcess = NULL,
        extraArgs  = NULL
    )
)
```

**Note:**

-   Here again, I initially tried many more models in addition to the
    above: `svmRadial`, `mlp`, `xgbTree`, `gamLoess`, `rocc`, `glmnet`,
    `vglmContRatio`, `RFlda`, and `lssvmRadial`, chosen arbitrarily from
    this [caret list](https://rdrr.io/cran/caret/man/models.html) - but
    they all underperformed by 30% or more, and the execution time was
    intolerable.

## Running the Models via Train

Here I write a function to generically execute a model via `train` and
capture its accuracy. Arguments allow for alternate data sets, formula,
model list, `trControl` and lists of `tuneGrid` objects.

```{r warning = FALSE, message = FALSE}
.train_model  <- function(
        model, cfg, .formula, .train, .test, bestTune,
       .trControl = NULL, .preProcess = NULL
) {
    tryCatch({
        fit <- (function(...) {
            caret::train(
                .formula,
                method     = model,
                data       = .train,
                trControl  = dplyr::coalesce(
                    .trControl,
                    trainControl(
                        method = 'repeatedcv',
                        number = 2,
                        p      = .5
                    )
                ),
                tuneGrid   = cfg$tuneGrid,
                preProcess = .preProcess,
                ...
            )
        })() # ideally, (...) 
    }, error = function(e) {
        print(c(Model = model, error = e))
    },
    finally = {
        if (exists('fit')) {
            if (bestTune) {
                print(c(Model = model, bestTune = fit$bestTune))
            }
            return(accuracy(fit, .test, type = 'raw'))
        }
    })
}

train_models <- function(
    .config       = train_config,
    .formula      = fc,
    .train        = train_df,
    .test         = test_df,
    bestTune      = FALSE,
    .trControl    = NULL,
    .preProcess   = NULL
) {
    results <- sapply(names(.config), function(model) {
        .train_model(
            model, .config[[model]], .formula, .train, .test,
            bestTune, .trControl, .preProcess
        )
    })
    names(results) <- names(.config)
    return(results)
}
```

**Notes:**

1.  The inner, anonymous function remains setup to accept extra
    arguments. I struggled mightily to send SOMETHING (key-value
    pair(s)) or NOTHING (i.e. *not* `NULL`). The only workaround so far
    is calling it with or without extra arguments via if/else, defeating
    the purpose of the ellipsis.
2.  I also got burned terribly here. Prior to wrapping in `tryCatch`,
    the function worked in most cases but died with some models for some
    parameters. Within `tryCatch`, I was getting nothing but `NA` - and
    a lot of them, even distilled down to one model. It seems `tryCatch`
    requires an explicit `return` statement, and that without it I was
    getting the value of the `finally` block...?

## First Runs - Baseline

And now I run what I'll consider a baseline.

```{r warning = FALSE, message = FALSE}
train_baselines <- train_models()
train_baselines |>
    as.data.frame() |>
    arrange(desc(baselines)) |>
    kable(col.names = c('Model', 'Accuracy')) |>
    kable_styling(full_width = FALSE, position = 'left')
```

Huh, `ranger` declined.

## Tuning the Models

Now I look for improvements via `tuneGrid`, `preProcess`, and
`trainControl` options.

### tuneGrid

I started with `tuneGrid` and ran the models once without any
customization and captured their out-of-the-box best `tuneGrid` values;
then repeated with various ranges around those values. This remains a
manual and tedious step, and generally not very effective. Below are the
values I settled upon after several iterations of nothing exciting.

```{r}
train_config <- list(
    ranger = list(
        tuneGrid = expand.grid(
            mtry          = 3,
            splitrule     = 'extratrees',
            min.node.size = 2
        )
    ),
    Rborist = list(
        tuneGrid = expand.grid(
            predFixed = 2,
            minNode   = 3
        )
    ),
    treebag = list(
    )
)

bestTunes <- train_models() #bestTune=TRUE)
bind_rows(baselines, train_baselines, bestTunes) |>
    as.data.frame() |>
    kable() |>
    kable_styling(full_width = FALSE, position = 'left') |>
    pack_rows(
        index = c('baseline' = 1, 'train() baseline' = 1, 'bestTune' = 1),
        label_row_css = 'text-align:center; background-color: #666; color: #fff;'
    )
```

This is terribly inconsistent, despite setting seed. Ironically,
`treebag` tends to change most, in either direction, despite *not*
having a `tuneGrid`. At least once, `ranger` plummeted 26% despite its
`tuneGrid` permutations including the values of the initial `bestTune`!
At any rate, there is generally not enough movement to consider
worthwhile.

### trainControl

Here I nullify the `trainGrid` settings and look at various
`trainControl` methods.

```{r warning = FALSE, message = FALSE}
train_config <- list(
    ranger = list(
        tuneGrid = NULL
    ),
    Rborist = list(
        tuneGrid = NULL
    ),
    treebag = list(
        tuneGrid = NULL
    )
)

boot_results <- bind_rows(
    train_models(.trControl = trainControl(
        method = 'boot', number = 2)),
    train_models(.trControl = trainControl(
        method = 'boot632', number = 2)),
    train_models(.trControl = trainControl(
        method = 'optimism_boot', number = 2)),
    train_models(.trControl = trainControl(
        method = 'boot_all', number = 2))
)

cv_results <- bind_rows(
    train_models(.trControl = trainControl(
        method = 'cv', p = .5, repeats = 2, number = 2)),
    train_models(.trControl = trainControl(
        method = 'repeatedcv', p = .5, repeats = 2, number = 2)),
    train_models(.trControl = trainControl(
        method = 'LGOCV', p = .5, repeats = 2, number = 2))
)

train_summary <- bind_cols(
    c(
        'boot', 'boot632', 'optimism_boot', 'boot_all',
        'cv', 'repeatedcv', 'LGOCV'
    ),
    bind_rows(boot_results, cv_results)
)

display_summary(train_summary)
```

**Note:**

-   There are also 'adaptive_boot', 'adaptive_cv', and 'adaptive_LGOCV'.
    They produced errors - *adaptive\$min should be less than 2*, and
    then *adaptive\$min should be greater than 1*, and finally *For
    adaptive resampling, there needs to be more than one tuning
    parameter for evaluation*, with `adaptive = list(min = 2, ...)`,
    `min = 1`, and `min = 1.1`. Infuriatingly, running the same on the
    command line worked. But for the good, this inspired `tryCatch`.

Conclusion - Again, not enough movement to consider worthwhile.

### preProcess

Now I try various `preProcess` methods. I do not try `scale` or
`center`, which I've already simulated; nor `cor` since I already
omitted `content`, half of the strongest correlating pair.

```{r warning = FALSE, message = FALSE}
pre_results <- bind_rows(
    train_models(.preProcess = c('BoxCox')),
    train_models(.preProcess = c('YeoJohnson')),
    train_models(.preProcess = c('expoTrans')),
    train_models(.preProcess = c('spatialSign')),
    train_models(.preProcess = c('pca'))
)

pre_summary <- bind_cols(
    c('BoxCox', 'YeoJohnson', 'expoTrans', 'spatialSign', 'pca'),
    pre_results
)

display_summary(pre_summary)
```

**Notes:**

There are numerous other `preProcess` methods:

-   I ruled out `zv` and `nzv`, as any column here with little variance
    is expected.

-   `knnImpute`, `bagImpute`, and `medianImpute` are not applicable as
    there are no missing values.

-   I looked further into `conditionalX`. Its description in
    `preProcess` help left me uncertain, and I found a fuller
    description alongside
    [nearZeroVar](https://search.r-project.org/CRAN/refmans/caret/html/nearZeroVar.html).
    In summary, it looks at the distribution of each predictor
    conditioned on the levels of the outcome, and identifies any
    predictor that is sparse within groups of the outcome.

    I ran `checkConditionalX` as follows - and no predictor was
    identified. (I will add this to my early analysis checklist, along
    with outliers, NAs, duplicates, etc .)

    ```{r}
    checkConditionalX(train_df |> select(-c(sentiment)), train_df$sentiment)
    ```

# Final Validation

The three models were close enough. Only `ranger` improved from
baseline, about equally for `preProcess` and `trainControl`.
`randomForest` and `treebag` moved so slightly I would discount it to
randomness.

```{r warning = FALSE, message = FALSE}
train_config <- list(
    ranger = list(
        tuneGrid = NULL
    ),
    Rborist = list(
        tuneGrid = NULL
    ),
    treebag = list(
        tuneGrid = NULL
    )
)

final_results <- train_models(
    .train = train_org,
    .test  = test_org
)

final_summary <- bind_cols(
    c('baseline', 'bestTune', 'trainControl', 'preProcess', 'finalValidation'),
    bind_rows(
        baselines,
        bestTunes,
        c(
            ranger  = max(train_summary[['ranger']]),
            Rborist = max(train_summary[['Rborist']]),
            treebag = max(train_summary[['treebag']])
        ),
        c(
            ranger  = max(pre_results[['ranger']]),
            Rborist = max(pre_results[['Rborist']]),
            treebag = max(pre_results[['treebag']])
        ),
        final_results
    )
)

display_summary(final_summary)
```

# Conclusions

1.  The winner is inconclusive; the numbers changes per run. (Did I NOT
    set seed?) I only see Python solutions on Kaggle, and they generally
    have train accuracy in the mid-90s and validation accuracy in the
    mid-80s. This Rmd matches that, albeit erratically...
2.  My primary goal was continued learning:
    -   Paragraph-based sentiments was the big lesson.
    -   I gained more comfort with `tryCatch` and `...`, but both came
        with surprises. I need to learn more, especially whether there
        is an idiomatic method of, for example, reading *optional*
        parameters/values from a config file and passing to an ellipses.
        Perhaps it is just that simple, but my experience here says
        otherwise.
    -   I had a note to myself to try
        [dtree](https://www.rdocumentation.org/packages/dtree/versions/0.4.2/topics/dtree)
        but when trying to install it I see that *Package ‘dtree’ was
        removed from the CRAN repository.*
3.  I had *many* troubles with this project - in addition to lengthy
    execution times against the full training set:
    -   Once, transitioning between Visual and Source editing modes, the
        closing backticks were removed from almost every code block,
        preventing Visual mode.
    -   Many times, a block of code failed or ran extremely slowly in
        RStudio but fine in the console - *or vice versa*.
    -   Once, manually running code blocks from the top down, each
        kickoff scrolled the document back to the top; forcing me to
        scroll further and further down to run the next block. What is
        that?!
    -   Already mentioned, accuracy changed wildly without plausible
        explanation, despite seed. Against the partitioned halves,
        accuracy fluctuated from the mid-60s to mid-90s?!
    -   I very often had to run code blocks *twice* in this Rmd: Often,
        a spurious second box of (empty) output is created, and
        disappears on subsequent run. Occasionally, the desired output
        is cropped and unscrollable - and self-corrects on a subsequent
        run. Occasionally, there is no output for no apparent reason -
        in which case I can often run just the output-producing line(s),
        proving that everything else worked.
4.  Once again, most effort was for naught, other than learning. The
    only thing that helped in a measurable way was the one obvious
    need - calculating sentiments.

# References

<https://rpubs.com/chelseyhill/676279>

<https://search.r-project.org/CRAN/refmans/caret/html/nearZeroVar.html>

<https://cran.r-project.org/web/packages/ranger/ranger.pdf>

<https://github.com/mjockers/syuzhet/blob/master/R/syuzhet.R>

# Appendix

## RandomForest Decision Tree

I like to end with a look at a model's decision tree. I've never seen
one with such symmetry! I can imagine the difficult choices.

```{r fig.width = 14, fig.height = 14}
fit <- randomForest(
    factor(sentiment) ~ .,
    train_df,
    ntree = 20
)   
simplest_tree <- which.min(fit$forest$ndbigtree)
fit$forest$xbestsplit <- round(fit$forest$xbestsplit, 2)  # 'digits' workaround
plot.getTree(
    fit,
    k     = simplest_tree,
    depth = 6,
    main  = sprintf(
        'RandomForest Decision Tree - Accuracy %.2f',
        accuracy(fit, test_df)
    )
)
```

## Case-Specific Random Forests

Reading further about `ranger`, I noticed `csrf` and thought I'd try it.
Alas, running against the full train/validation sets gives, *Error in
table(data, reference, dnn = dnn, ...) : all arguments must have the
same length*. That seems odd - is it uncommon to have different sized
data sets...?

```{r}
pred <- csrf(
    factor(sentiment) ~ .,
    training_data = train_df,
    test_data     = test_df,
    params1       = list(num.trees = 50, mtry = 4),
    params2       = list(num.trees = 5)
)

cm <- confusionMatrix(pred, factor(test_df$sentiment))
cm$overall[['Accuracy']]
```
