---
title: "Kaggle Plant Disease Classification"
author: "hbmaclean"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
editor_options:
  markdown:
    toc: true
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{css, echo = FALSE}
h1 {
    margin-top: 2em;
    border-bottom: 2px solid black;
    padding-bottom: 5px;
}
h2 {
    margin-top: 1em;
}
h3 {
    margin-top: 1em;
}
p {
    margin-top: 2px;
}
table {
    margin-top: 1em;
    margin-bottom: 1em;
    background: snow;
}
th {
    background: snow;
}
img {
    border: 1px solid black;
    margin-top: 4px;
    margin-bottom: 4px;
    width: 100%;
}
```

# Introduction {#introduction}

This is just another side project to keep me learning. I picked this one
in particular to try a source other than
[UCI](https://archive.ics.uci.edu/datasets) and because it had some
notebooks listed that I can gauge against. I scanned about five of the
notebooks - all were python-based, looked very similar, and had accuracy
around 75-85%.

Below is a description of the data from
[Kaggle](https://www.kaggle.com/datasets/turakut/plant-disease-classification).

> Plant diseases cause significant agricultural losses worldwide. Early
> prediction of disease outbreaks can help farmers take preventive
> measures. This synthetic dataset simulates environmental conditions
> that might lead to fungal infections in plants.
>
> The dataset contains 10,000 samples representing environmental
> measurements from different farm locations with the following
> features: 
>
> -   **temperature**: Measured in degrees Celsius 
>
> -   **humidity**: Measured as percentage 
>
> -   **rainfall**: Measured in millimeters 
>
> -   **soil_pH**: Acidity/alkalinity measurement 
>
> -   **disease_present**: Binary label (0 = healthy, 1 = diseased)

# Importing Packages

```{r warning = FALSE, message = FALSE}
repo <- "http://cran.us.r-project.org"
packages <- c(
    'curl', 'stringr', 'dplyr', 'tidyr', 'data.table', 'caret', 'tibble',
    'kableExtra', 'ggcorrplot', 'gridExtra', 'GGally', 'stats', 'rstatix',
    'ranger', 'tuneRanger', 'mlr', 'randomForest', 'missMethods', 'nnet',
    'naivebayes', 'matrixStats', 'earth', 'mda', 'hda', 'bnclassify', 'rocc',
    'VGAM', 'HiDimDA', 'DT', 'devtools'
)
for (package in packages) {
    if (!require(package, character.only = TRUE)) {
        install.packages(package, repos = repo)
        library(package, character.only = TRUE)
    }
}

install_github('munoztd0/reprtree')
eval(substitute(library(pkg), list(pkg='reprtree')))
library(reprtree)

set.seed(34)
```

# Fetching the Data

Rather than setting up API keys and assuming the reader will do the
same, I fetch instead with `curl`. Straightforward, other than that the
CSV has column names in the first row.

```{r}
fetch_data <- function() {
    options(timeout = 120)
    url <- paste0(
        'https://www.kaggle.com/api/v1/datasets/download/',
        'turakut/plant-disease-classification'
    )
    tmp <- tempfile()
    curl_download(url, tmp)
    data <- as.data.frame(
        read.csv(
            unz(tmp, 'plant_disease_dataset.csv'),
            header = FALSE,   # first row has column names
            skip = 1          # skip them
        )
    )
    colnames(data) <- c(
        'temperature', 'humidity', 'rainfall', 'soil_ph', 'disease_present'
    )
    return(data)
}
```

## Data Sample

```{r}
plant_df <- fetch_data()
head(plant_df) |> kable() |> kable_styling()
```

# Partitioning

```{r}
index    <- createDataPartition(y = plant_df$disease_present, times = 1, p = 0.5, list = FALSE)
train_df <- plant_df[index,]
test_df  <- plant_df[-index,]
```

## Class Distribution {#class_distribution}

Good for the plants - only 24% are diseased.

```{r}
prop.table(table(plant_df$disease_present)) |>
    kable(col.names = c('disease_present', 'proportion')) |>
    kable_styling(full_width = FALSE, position = 'left')
```

Here I look at histograms of the predictors faceted by
`disease_present`. To my eye, each predictor follows one very similar
distribution, whether healthy or diseased. The y-axes indicate the class
imbalance.

```{r warning = FALSE, fig.width=10, fig.height=10}
common <- c(
     geom_histogram(binwidth = 5),
     facet_wrap(
         disease_present ~ .,
         nrow     = 2,
         scales   = 'free',
         labeller = as_labeller(c('0' = 'Healthy', '1' = 'Diseased'))
     )
)

p1 <- plant_df |> ggplot(aes(humidity))    + common
p2 <- plant_df |> ggplot(aes(rainfall))    + common
p3 <- plant_df |> ggplot(aes(temperature)) + common
p4 <- plant_df |> ggplot(aes(soil_ph))     + common
grid.arrange(
    p1, p2, p3, p4, ncol = 2
)
```

## Missing Values, Outliers, Extreme Values {#missing_outlying_extreme}

There are no missing values nor extreme values, but evidently some
outliers to potentially impute.

```{r}
c(
    'missing'  = any(is.na(plant_df)),
    'outlying' = any(rstatix::is_outlier(plant_df |> as.matrix())),
    'extreme'  = any(rstatix::is_extreme(plant_df |> as.matrix()))
) |> kable(col.names = c('Values', 'Occurrence')) |>
     kable_styling(full_width = FALSE, position = 'left')
```

Since there are outliers, I'll tuck aside a copy of `train_df` with
outliers set to `NA` - to be imputed later.

```{r}
train_o      <- train_df |> as.matrix()
outliers     <- rstatix::is_outlier(train_o)
idx          <- which(outliers[,2:ncol(train_o)])
train_o[idx] <- NA
train_o      <- train_o |> as.data.frame()
```

# Row- and Column SDs

Here I use standard deviation as a proxy for volatility, and look for an
observable difference between diseased and healthy cases.

```{r}
x  <- plant_df$disease_present
y  <- matrixStats::rowSds(plant_df[,1:4] |> as.matrix())
ggplot() +
    geom_boxplot(aes(x, y, group = x)) +
    scale_x_continuous(
        breaks = c(0,1),
        labels = as_labeller(c('0' = 'Healthy', '1' = 'Diseased'))
    ) +
    labs(x = '', y = 'rowSds')
```

Row-wise, diseased observations have a narrower IQR and considerably
higher mean.

Now I take a similar look at columns' standard deviations and means.

```{r warning = FALSE}
plant_df |>
    group_by(disease_present) |>
    dplyr::summarize(across(everything(), list(mean = mean, sd = sd))) |>
    kable(
        digits = 3,
        col.names = c('diseased', rep(c('mean', 'sd'), 4))
    ) |>
    kable_styling(
        bootstrap_options = c('bordered', 'striped'),
        full_width = FALSE,
        position = 'left'
    ) |>
    add_header_above(c(' ', 'temp' = 2, 'humidity' = 2, 'rain' = 2, 'soil/ph' = 2))
```

Column-wise, the two classes are fairly similar in terms of
`temperature` and `soil_ph`; but diseased observations have considerably
higher mean for `humidity` and `rainfall`, as well as a considerably
higher deviation for `rainfall`. Perhaps these facts can be incorporated
as case weights or formula interactions.

# Dimension Reduction

I have low expectations here since there are only four predictors.

## Near-Zero Variability

To the contrary, almost all data points are unique.

```{r}
plant_df |> select(-c(disease_present)) |>
    summarise_all(n_distinct) |>
    pivot_longer(everything(), names_to = 'Column', values_to = 'N_Distinct') |>
    mutate(Variability = N_Distinct/nrow(plant_df) * 100) |>
    arrange(desc(N_Distinct)) |>
    knitr::kable(
        digits  = 2,
        align   = 'l',
        caption = 'Plant Disease Feature Variability'
    ) |>
    kable_styling(
        bootstrap_options = c('bordered', 'striped'),
        full_width = FALSE,
        position = 'left'
    )
```

## Principal Component Analysis

Nothing worthwhile here either - all features contribute rather evenly
to the overall variance.

```{r}
pca <- prcomp(plant_df, scale = TRUE)
PCs <- which(cumsum(pca$sdev^2)/sum(pca$sdev^2) < 0.90)
s   <- summary(pca)
p1  <- barchart(
    pca$rotation[,1:length(PCs)],
    auto.key = TRUE,
    main = sprintf(
        '90%% variance PCs %s',
        paste(PCs, collapse = ', ')
    )
)
p2 <- s$importance[,1:4] |>
    as.data.frame() |>
    rownames_to_column() |>
    filter(rowname == 'Proportion of Variance') |>
    pivot_longer(-rowname) |>
    dplyr::select(-rowname) |>
    mutate(name = reorder(name, desc(value))) |>
    ggplot(aes(name, value)) +
    geom_col() +
    labs(x = '', y = 'Variance Explained', title = 'Scree Plot') +
    theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) +
    scale_y_continuous(labels = scales::percent)

grid.arrange(p1, p2, nrow = 2)
```

## Correlation

All correlations are weak, features appear to be independent of one
another.

```{r}
# matrix of correlation p-values (conflict with rstatix w/o qualification)
p.mat <- ggcorrplot::cor_pmat(plant_df)

ggcorrplot(
    cor(plant_df),
    hc.order    = TRUE,     # reorder layout with hierarchical clustering
    hc.method   = 'median', # see ?hclust for other methods
    lab         = TRUE,     # adds correlation coefficients
    lab_size    = 3,
    digits      = 1,
    show.diag   = FALSE,    # omits self-correlation diagonal
    p.mat       = p.mat,    # p-values
    sig.level   = 0.001,    # p-value threshold
    insig       = 'blank',  # omits correlations with p-values below threshold
    show.legend = FALSE,
    title       = 'Plant Feature Correlations'
)
```

But, `humidity`, `rainfall`, and `soil_ph` correlations with the outcome
have very low p-values, indicated by `***`.

```{r message = FALSE}
ggpairs(plant_df)
```

## Analysis of Variance (AOV)

[AOV](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/aov)
does caution that it works best with balanced designs - and I assume
this is NOT, given the class distribution of about 75%/25%. Otherwise, I
would conclude that `rainfall` and `humidity` are most influential.

```{r}
unclass(summary(stats::aov(disease_present ~ ., data = plant_df))) |>
    as.data.frame() |>
    rownames_to_column(var = 'Column') |>
    filter(!is.na(F.value)) |>
    mutate(Column = reorder(Column, desc(F.value))) |>
    ggplot(aes(Mean.Sq, F.value, color = Column)) +
    geom_point(cex = 4) +
    scale_x_log10()
```

Concluding this topic, there is no apparent room for reduction. `corr`,
`pca`, and `nzv` will surely not be worthwhile `preProcess` options.

# Initial Model Framework

Here I create functions for making predictions, calculating accuracy,
running individual models, tracking results, and easily running them all
together. The goal is to run numerous models and then narrow down to the
most accurate one or few.

## Predictions and Accuracy

Below are functions to make predictions and calculate accuracy, using
the
[ellipsis](https://cran.r-project.org/web/packages/ellipsis/ellipsis.pdf)
argument to allow model nuances to be passed in.

```{r}
predictions <- function(model, test_df, ...) {
    pred <- predict(model, test_df, ...)
    if (is.atomic(pred)) {
        predictions <- pred
    }
    else {
        predictions <- pred$predictions
    }
    predictions
}

accuracy <- function(model, test_df, verbose = FALSE, ...) {
    acc <- NA
    tryCatch({
        pred <- predictions(model, test_df, ...)
        cm   <- confusionMatrix(pred, factor(test_df$disease_present))
        acc  <- round(100*cm$overall[['Accuracy']], 3)
    },
    error = function(e) {
        if (verbose) {
            sprintf('model %s, error %s', class(model), e)
        }
    },
    finally = {
        return(acc)
    })
}
```

## Formula and Models

And here, a basic formula; and a function for each model.

```{r}
fc <- as.formula(factor(disease_present) ~ .)

ranger__c <- function(train_df, test_df, formula = fc) {
    fit <- ranger(formula, train_df)
    accuracy(fit, test_df)
}

rf__c <- function(train_df, test_df, formula = fc) {
    fit <- randomForest(formula, train_df, ntree = 20)
    accuracy(fit, test_df)
}

multinom__c <- function(train_df, test_df, formula = fc) {
    fit <- multinom(formula, train_df, trace = FALSE, decay = .00015)
    accuracy(fit, test_df)
}

naive_bayes__c <- function(train_df, test_df, formula = fc) {
    fit <- naive_bayes(formula, train_df, usekernel = TRUE)
    accuracy(fit, test_df)
}

avNNet__c <- function(train_df, test_df, formula = fc) {
    fit <- avNNet(formula, train_df, size = 5, trace = FALSE)
    accuracy(fit, test_df, type = 'class')
}

knn3__c <- function(train_df, test_df, formula = fc, ...) {
    fit <- knn3(formula, train_df, ...)
    accuracy(fit, test_df, type = 'class')
}
```

## Model Runner

Finally, a function to actually run the models and accumulate metrics.

```{r}
run_models <- function(.train = train_df, .test = test_df, ...) {
    results <- vector()
    results <- append(results, c(
        ranger__c(.train, .test, ...),
        rf__c(.train, .test, ...),
        multinom__c(.train, .test, ...),
        naive_bayes__c(.train, .test, ...),
        avNNet__c(.train, .test, ...),
        knn3__c(.train, .test, k = 13)
    ))
    names(results) <- c('ranger', 'rf', 'multinom', 'naive_bayes', 'avNNet', 'knn3')
    return(results)
}
```

## Metrics Summary

Below, a function to display the metrics. At the moment, the expectation
is that the named vector from `run_models` will be bound with an initial
column of meaningful labels.

```{r}
display_summary <- function(.data) {
    .data |>
        dplyr::rename(` ` = c(1)) |>
        datatable(
            rownames = FALSE,
            width    = '75%',
            options  = list(
                dom        = 't',
                pageLength = nrow(.data)
            )
        )
}
```

## Initial Runs - Baseline, Imputing, Scaling, Centering

With the framework defined, I run the models in various scenarios -
imputing outliers, scaling, and centering.

```{r warning = FALSE, message = FALSE}
model_summary <- bind_rows(
    run_models(),
    run_models(.train = impute_mean(train_o)),
    run_models(.train = impute_median(train_o)),
    run_models(
        .train = train_df |>
            mutate(across(-c(disease_present), ~ scale(.x, center=FALSE, scale=TRUE))),
        .test  = test_df  |>
            mutate(across(-c(disease_present), ~ scale(.x, center=FALSE, scale=TRUE)))
    ),
    run_models(
        .train = train_df |>
            mutate(across(-c(disease_present), ~ scale(.x, center=TRUE, scale=FALSE))),
        .test  = test_df  |>
            mutate(across(-c(disease_present), ~ scale(.x, center=TRUE, scale=FALSE)))
    ),
    run_models(
        .train = train_df |>
            mutate(across(-c(disease_present), ~ scale(.x, center=TRUE, scale=TRUE))),
        .test  = test_df  |>
            mutate(across(-c(disease_present), ~ scale(.x, center=TRUE, scale=TRUE)))
    )
)
model_summary <- bind_cols(
    c(
        'baseline', 'mean-impute outliers', 'median-impute outliers',
        'scaled', 'centered', 'centered & scaled'
    ),
    model_summary
)

display_summary(model_summary)
```

# Second Model Framework - train()

From above, `ranger` and `randomForest` look best - always above 80%.
Here, I look more closely at those two using `train`, hoping to tune
them better with `preProcess`, `tuneGrid`, and `trainControl`.

## Models and Tuning

Many models seem to be available only via `train`; others are available
directly but do not conform to the typical fit/predict/confusionMatrix
pattern. Since I am building a framework to run `ranger` and
`randomForest` via `train`, I also chose additional classification
models that I have used before or sampled arbitrarily now from this
[caret list](https://rdrr.io/cran/caret/man/models.html).

First, I define a list of `tuneGrid` objects for the models, each `NULL`
by default; with the names serving as a proxy for the list of models.

```{r}
tuning  <- list(
    'ranger'        = NULL,
    'rf'            = NULL,
    'Rborist'       = NULL,
    'svmRadial'     = NULL,
    'mlp'           = NULL,
    'xgbTree'       = NULL,
    'gamLoess'      = NULL,
    'rocc'          = NULL,
   #'hda'           = NULL,   <---- omitted, hopelessly verbose
    'glmnet'        = NULL,
    'treebag'       = NULL,
    'vglmContRatio' = NULL,
    'RFlda'         = NULL,
    'lssvmRadial'   = NULL
)
```

This is admittedly a shotgun approach, but it exposes me to more models.
I actually chose more models, but skipped any that failed for any
reason. Some require packages that are not in the usual location; and
the following failed with, "all the Accuracy metric values are missing":

-   gbm, fda, awnb, bagFDAGCV, ada, rpartCost, dwdRadial, randomGLM,
    binda, lssvmLinear

I have encountered the error before and I suspect that either it could
be resolved with appropriate tuning, or the model is simply not designed
for the data. I may try to solve that some other day.

## Running the Models via Train

Here I write a function to generically execute a model via `train` and
capture its accuracy. Arguments allow for alternate data sets, formula,
model list, `trControl` and lists of `tuneGrid` objects; and the
[ellipsis](https://cran.r-project.org/web/packages/ellipsis/ellipsis.pdf)
argument implicitly allows for `preProcess` and whatever else might be
applicable.

```{r warning = FALSE, message = FALSE}
execute_models <- function(
    .train        = train_df,
    .test         = test_df,
    .formula      = fc,
    .models       = names(tuning),
    .tuning       = tuning[c(.models)],
    .trControl    = trainControl(method = 'repeatedcv', number = 2, p = .5),
    .preProcess   = c('center', 'scale'),
    bestTune      = FALSE,
    verbose       = FALSE,
    ...
) {
    results <- vector()
    results <- mapply(
        function(model, tune, ...) {
            acc <- NA
            tryCatch({
                fit <- caret::train(
                      .formula,
                      method        = model,
                      data          = .train,
                      trControl     = .trControl,
                      tuneGrid      = tune,
                      preProcess    = .preProcess,
                      trace         = FALSE,
                      verbosity     = 0,
                      verbose       = FALSE,
                      na.action     = na.pass,
                      ...
                )
                acc <- accuracy(fit, .test)
            },
            error = function(e) {
                if (verbose) {
                    print(c(Model = model, error = e))
                }
            },
            finally = {
                results <- append(results, acc)
                if (bestTune && !is.null(acc)) {
                    print(c(Model = model, bestTune = fit$bestTune))
                }
            })
        },
        .models,
        .tuning,
        ...
    )
    if (is.vector(results) & length(names(results))>1) {
        return(bind_rows(results))
    }
    else {
        return(results)
    }
}
```

## First Runs - Baseline

And now I run what I'll consider a baseline and look for the top few.

```{r warning = FALSE, message = FALSE}
baselines <- execute_models()
baselines |>
    pivot_longer(everything()) |>
    arrange(desc(value)) |>
    kable(col.names = c('Model', 'Accuracy')) |>
    kable_styling(full_width = FALSE, position = 'left')
```

## Tuning the Top Five Models

Next I'll focus on the top five models, which appropriately are all
forestry-related in name. First I will dig into each model's `tundGrid`.

It proved to be a little tricky passing in the `tuneGrid` for a single
model. Since I've implemented `.tuning` as an argument to `mapply`, a
list is expected. Passing in a single `tuneGrid` gives an error,
"incorrect number of dimensions".

Below is one example; and for `xbgTree` I had to be careful to NOT
create too large a grid!

```{r eval = FALSE}
execute_models(
    .models = c('xgbTree'),
    .tuning = list(expand.grid(
        nrounds          = 10,
        max_depth        = c(2:5),
        eta              = c(.5,.7,.9),
        gamma            = c(3:6),
        colsample_bytree = c(.7,.8,.9),
        min_child_weight = c(5,10,15),
        subsample        = 1
    )),
    bestTune = TRUE
)
```

Below are the settings I settled on. Fiddling them further did not
change accuracy much in either direction.

```{r warning = FALSE}
tuning <- list(
    xgbTree    = expand.grid(
        nrounds          = 50,
        max_depth        = 3,
        eta              = .1,
        gamma            = .9,
        colsample_bytree = seq(.8,1,.1),
        min_child_weight = c(1:3),
        subsample        = c(1:2)
    ),
    ranger     = data.frame(
        mtry          = 2,
        splitrule     = 'hellinger',
        min.node.size = 50
    ), 
    rf      = data.frame(mtry = 2),
    Rborist = data.frame(predFixed = 2, minNode   = 2),
    treebag = NULL
)
tuned <- execute_models()
tuned |>
    pivot_longer(everything()) |>
    arrange(desc(value)) |>
    kable(col.names = c('Model', 'Accuracy')) |>
    kable_styling(full_width = FALSE, position = 'left')
```

## Pre-Processing the Top Five Models

With the top five models tuned, I look into other `preProcess` methods.

Note, "all the Accuracy metric values are missing" occurs again with the
following methods:

-   'conditionalX'
-   'knnImpute'
-   'BoxCox', but only when adjusting the associated 'fudge' parameter,
    and then only for 'ranger' and 'rf'

```{r warning = FALSE, message = FALSE}
preprocessed <- bind_rows(
    execute_models(.preProcess = c('BoxCox')),
    execute_models(.preProcess = c('YeoJohnson')),
    execute_models(.preProcess = c('expoTrans')),
    execute_models(.preProcess = c('center')),
    execute_models(.preProcess = c('scale')),
    execute_models(.preProcess = c('center', 'scale')),
    execute_models(.preProcess = c('bagImpute'), .train = train_o),
    execute_models(.preProcess = c('medianImpute'), .train = train_o)
)
```

And the results:

```{r warning = FALSE, message = FALSE}
model_summary <- bind_cols(
    c(
        'baseline', 'tuned',
        'BoxCox', 'YeoJohnson', 'expoTrans', 'center', 'scale',
        'center+scale', 'bagImpute', 'medianImpute'
    ),
    bind_rows(baselines[c(names(tuned))], tuned, preprocessed),
)

display_summary(model_summary)
```

Conclusions from this:

-   as seen earlier projects, imputing outliers has a measurably
    *harmful* effect
-   otherwise, various `preProcess` options made *minor* improvements
    over the 'tuned' models
-   `treebag` is lagging - its best accuracy is less than the others'
    least, other than imputing

## trainControl Methods for the Top Five Models

Now I try various `trainControl` methods. For simplicity, I rely mostly
on underlying default values; and specify only `method`, and constant
`number = 2` rather than default 10 or 25, for performance reasons.

```{r warning = FALSE, message = FALSE}
# Note, there is also LOOCV (leave-one-out CV) but it is computationally *intense*
trcontrolled <- bind_rows(
    execute_models(.trControl = trainControl(number = 2, method = 'boot')),
    execute_models(.trControl = trainControl(number = 2, method = 'boot632')),
    execute_models(.trControl = trainControl(number = 2, method = 'boot_all')),
    execute_models(.trControl = trainControl(number = 2, method = 'optimism_boot')),
    execute_models(.trControl = trainControl(method = 'adaptive_boot')),
    execute_models(.trControl = trainControl(number = 2, method = 'cv')),
    execute_models(.trControl = trainControl(number = 2, method = 'repeatedcv')),
    execute_models(.trControl = trainControl(method = 'adaptive_cv')),
    execute_models(.trControl = trainControl(number = 2, method = 'LGOCV')),
    execute_models(.trControl = trainControl(method = 'adaptive_LGOCV'))
)
trcontrolled <- bind_cols(
    c(
        'boot', 'boot632', 'boot_all', 'optimism_boot', 'adaptive_boot',
        'cv', 'repeatedcv', 'adaptive_cv', 'LGOCV', 'adaptive_LGOCV'
    ),
    trcontrolled
)

model_summary <- bind_rows(model_summary, trcontrolled)
display_summary(model_summary)
```

## Formula Modifications

I understand the *syntax* of what I'm doing here, but I cannot envision
the underlying mechanics. I have no expectations, just another
experiment.

```{r warning = FALSE, message = FALSE}
formulamods <- bind_rows(
    execute_models(
        .formula = as.formula(factor(disease_present) ~ humidity*rainfall + .)
    ),
    execute_models(
        .formula = as.formula(
            factor(disease_present) ~ humidity:rainfall + .
        )
    ),
    execute_models(
        .formula = as.formula(
            factor(disease_present) ~ (humidity+rainfall+temperature)^3 + .
        )
    ),
    execute_models(
        .formula = as.formula(
            factor(disease_present) ~ (humidity+rainfall+temperature)^3 + I(soil_ph^2)
        )
    )
)

formulamods <- bind_cols(
    c(
        'humidity*rain', 'humidity:rain', '(humidity+rain+temp)^3',
        '(humidity+rain+temp)^3 + ph^2'
    ),
    formulamods
)

model_summary <- bind_rows(model_summary, formulamods)
display_summary(model_summary)
```

That had a negative effect, though not quite as bad as imputing.

Note, I tried similar formulas when running the models directly (without
`train`) but continually failed with, "One or more independent variables
not found in data."

# Final Efforts with Top Two

## Combinations of Isolated Improvements

Looking at the running summary, some settings *appeared* to make minor
improvements beyond the initial tuning, in isolation. I certainly do not
assume that combining such improvements will result in a combined
improvement - but I will try some combinations, for the top two models.

The following settings showed some promise for `xgbTree` and `ranger`,
all somewhat above baseline accuracy. The table is not meant to convey
combinations; the two columns are independent, each setting was in
isolation.

```{r echo = FALSE}
tribble(
    ~trainControl, ~preProcess, 
    'adaptive_cv', 'center, scale',
    'optimism_boot', 'center',
    'LGOCV', '',
    'cv', '',
    'adaptive_LGOCV', '',
    'repeatedcv', 'center',
    'LGOCV', 'scale',
    'cv', 'BoxCox'
) |>
    kable() |>
    kable_styling(
        bootstrap_options = c('bordered'),
        full_width = FALSE,
        position = 'left'
    ) |>
    pack_rows(
        index = c('xgbTree' = 5, 'ranger' = 3),
        label_row_css = 'text-align:center; background-color: #666; color: #fff;'
    )
```

### xgbTree

First, I run the `xgbTree` with many permutations of `trainControl` and
`preProcess` settings shown above.

```{r warning = FALSE}
.xgbTree <- bind_rows(
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox'),
        .trControl  = trainControl(method = 'adaptive_cv')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('center'),
        .trControl  = trainControl(method = 'adaptive_cv')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('scale'),
        .trControl  = trainControl(method = 'adaptive_cv')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox', 'center', 'scale'),
        .trControl  = trainControl(method = 'adaptive_cv')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox'),
        .trControl  = trainControl(method = 'optimism_boot')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('center'),
        .trControl  = trainControl(method = 'optimism_boot')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('scale'),
        .trControl  = trainControl(method = 'optimism_boot')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox', 'center', 'scale'),
        .trControl  = trainControl(method = 'optimism_boot')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox'),
        .trControl  = trainControl(method = 'LGOCV')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('center'),
        .trControl  = trainControl(method = 'LGOCV')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('scale'),
        .trControl  = trainControl(method = 'LGOCV')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox', 'center', 'scale'),
        .trControl  = trainControl(method = 'LGOCV')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox'),
        .trControl  = trainControl(method = 'cv')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('center'),
        .trControl  = trainControl(method = 'cv')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('scale'),
        .trControl  = trainControl(method = 'cv')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox', 'center', 'scale'),
        .trControl  = trainControl(method = 'cv')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox'),
        .trControl  = trainControl(method = 'adaptive_LGOCV')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('center'),
        .trControl  = trainControl(method = 'adaptive_LGOCV')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('scale'),
        .trControl  = trainControl(method = 'adaptive_LGOCV')
    ),
    execute_models(
        .models     = c('xgbTree'),
        .preProcess = c('BoxCox', 'center', 'scale'),
        .trControl  = trainControl(method = 'adaptive_LGOCV')
    )
)
```

And the results:

```{r message = FALSE}
bind_cols(
        c(
            rep('adaptive_cv', 4),
            rep('optimism_boot', 4),
            rep('LGOCV', 4),
            rep('cv', 4),
            rep('adaptive_LGOCV', 4)
        ),
        c(
            rep(c('BoxCox', 'center', 'scale', 'BoxCox,center,scale'), 5)
        ),
       .xgbTree
) |> 
    dplyr::rename(
        trControl = c(1),
        preProcess = c(2),
        Accuracy = c(3)
    ) |>
    datatable(
        rownames = FALSE,
        width    = '50%',
        caption  = 'xgbTree',
        options  = list(dom = 't', pageLength = nrow(.xgbTree))
    )
```

### ranger

Now the same for `ranger`.

```{r warning = FALSE}
.ranger <- bind_rows(
    execute_models(
        .models = c('ranger'),
        .preProcess = c('BoxCox'),
        .trControl = trainControl(method = 'repeatedcv')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('center'),
        .trControl = trainControl(method = 'repeatedcv')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('scale'),
        .trControl = trainControl(method = 'repeatedcv')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('BoxCox', 'center', 'scale'),
        .trControl = trainControl(method = 'repeatedcv')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('BoxCox'),
        .trControl = trainControl(method = 'LGOCV')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('center'),
        .trControl = trainControl(method = 'LGOCV')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('scale'),
        .trControl = trainControl(method = 'LGOCV')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('BoxCox', 'center', 'scale'),
        .trControl = trainControl(method = 'LGOCV')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('BoxCox'),
        .trControl = trainControl(method = 'cv')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('center'),
        .trControl = trainControl(method = 'cv')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('scale'),
        .trControl = trainControl(method = 'cv')
    ),
    execute_models(
        .models = c('ranger'),
        .preProcess = c('BoxCox', 'center', 'scale'),
        .trControl = trainControl(method = 'cv')
    )
)
```

And the results:

```{r message = FALSE}
bind_cols(
        c(
            rep('repeatedcv', 4),
            rep('LGOCV', 4),
            rep('cv', 4)
        ),
        c(
            rep(c('BoxCox', 'center', 'scale', 'BoxCox,center,scale'), 3)
        ),
       .ranger
) |> 
    dplyr::rename(
        trControl = c(1),
        preProcess = c(2),
        Accuracy = c(3)
    ) |>
    datatable(
        rownames = FALSE,
        width    = '50%',
        caption  = 'ranger',
        options  = list(dom = 't', pageLength = nrow(.ranger))
    )
```

I will stop there, having turned every dial I can conceive.

# Loose Ends

## Ranger Case- and Class Weights

`ranger` supports case- and class weights. I have used them before and
know they can be very effective at balancing lopsided data. So far I've
found the balancing to be a zero-sum effort in terms of accuracy/RMSE;
for example, accuracy for true negatives may improve at the expense of
true positives.

I create case weights (weighing each observation) with
`parsnip::importance_weights` and, just a guess, `rowSds`.

To weigh the two classes, I refer again to [this
article](https://medium.com/@ravi.abhinav4/improving-class-imbalance-with-class-weights-in-machine-learning-af072fdd4aa4).

```{r}
.case_weights = parsnip::importance_weights(
    rowSds(train_df[1:4] |> as.matrix()) |>
    as.vector()
)
.class_weights <- c(
    nrow(train_df) / (2 * sum(train_df$disease_present == 0)),
    nrow(train_df) / (2 * sum(train_df$disease_present == 1))
)
```

I know from the past that weights may work seamlessly via `train` for
*bootstrap methods*; but fail for cross-validation methods. This makes
sense, as each cross-validation fold will be shorter in length than the
full list of weights. I use a known workaround for cv, after trying
`train` and bootstrap.

### Weights with Bootstrap via Train

```{r warning = FALSE}
.boot <- bind_rows(
      execute_models(
          .models        = c('ranger'),
          .trControl     = trainControl(method = 'boot'),
          .preProcess    = c('center')
      ),
      execute_models(
          .models        = c('ranger'),
          .trControl     = trainControl(method = 'boot'),
          .preProcess    = c('center'),
          case.weights   = list(.case_weights)
      ),
      execute_models(
          .models        = c('ranger'),
          .trControl     = trainControl(method = 'boot'),
          .preProcess    = c('center'),
          class.weights  = list(.class_weights)
      ),
      execute_models(
          .models        = c('ranger'),
          .trControl     = trainControl(method = 'boot'),
          .preProcess    = c('center'),
          case.weights   = list(.case_weights),
          class.weights  = list(.class_weights)
      )
)
```

### Weights with Cross-Validation

Here I simulate cross-validation per [this
solution](https://github.com/imbs-hl/ranger/issues/419), iterating over
10 folds. It begs the question though, given one accuracy result per
fold, what is the final accuracy? Perhaps average?

```{r}
n            <- nrow(train_df)
num_folds    <- 10
folds        <- split(sample(n), cut(1:n, num_folds, labels = FALSE))

.cv <- sapply(folds, function(fold) {
    .test  <- train_df[fold,]
    .train <- train_df[-fold,]
    fit    <- ranger(factor(disease_present) ~ ., .train)
    fit_a  <- ranger(factor(disease_present) ~ ., .train, case.weights = .case_weights[-fold])
    fit_b  <- ranger(factor(disease_present) ~ ., .train, class.weights = .class_weights)
    fit_ab <- ranger(factor(disease_present) ~ ., .train, case.weights = .case_weights[-fold], class.weights = .class_weights)
    c(
        baseline     = accuracy(fit, .test),
        caseWeights  = accuracy(fit_a, .test),
        classWeights = accuracy(fit_b, .test),
        bothWeights  = accuracy(fit_ab, .test)
    )
})
```

And the results:

```{r message = FALSE}
bind_cols(
    c('baseline', 'caseWeights', 'classWeights', 'bothWeights'),
    .cv,
    .boot
) |> kable(col.names = c('', paste(rep('fold',10), 1:10), '')) |>
     add_header_above(c('', 'cross-validation' = 10, 'boot' = 1)) |>
     kable_styling(
         bootstrap_options = c('bordered', 'striped'),
         full_width = FALSE,
         position = 'left'
    )
```

I conclude that neither class- nor case weights help accuracy here.
(Granted, my calculation for case weights was guesswork.)

I leave open the question whether weights helped balance specificity and
sensitivity.

### Weights with dplyr

[This article](https://www.tidyverse.org/blog/2022/05/case-weights/)
suggests another approach, which I will try here.

```{r warning = FALSE}
execute_models(
    .models        = c('ranger'),
    .trControl     = trainControl(method = 'cv'),
    .preProcess    = c('center'),
    .train         = train_df |> dplyr::mutate(
        case_wts = parsnip::importance_weights(.class_weights[disease_present+1])
    ),
    .test          = test_df  |> dplyr::mutate(
        case_wts = parsnip::importance_weights(.class_weights[disease_present+1])
    )
) |>
    kable(col.names = c('', 'Accuracy?!')) |>
    kable_styling(bootstrap_options = 'bordered', full_width = FALSE, position = 'left')
```

Hold on, this *must* be cheating - it creates another binary column
having 100% correlation with the outcome...

This seems equivalent to setting `case_wts = disease_present`, or adding
the outcome to the right hand side of the formula, e.g.
`factor(disease_present) ~ . + disease_present` (both of which also give
100% accuracy.)

## Custom-Tuning Ranger

While poking around, I found
[tuneRanger](https://cran.r-project.org/web/packages/tuneRanger/tuneRanger.pdf).
Its `tuneMtryFast` indicates that the out-of-bag (OOB) error varies very
little for any `mtry` near 1. The plot looks dramatic but note the tiny
range on the y-axis. (I would not have guessed that `mtry` could be zero
or fractional.)

```{r results = FALSE, message = FALSE, error = FALSE, warning = FALSE}
sink(nullfile())
res <- tuneMtryFast(
    fc,
    data       = train_df,
    mtryStart  = .1,
    stepFactor = .1,
    improve    = .05,
    trace      = FALSE,
    plot       = FALSE
)
sink()
plot(res, type = 'b')
```

### Comparing tuneRanger/ranger and Positive Class

The package also provides some other tools worth trying. Here I try
`tuneRanger`, which first needs a "task" object from `mlr`. Then,
evidently, `estimateTimeTuneRanger` will estimate how long to expect to
process the "task" with `tuneRanger`; and a prediction can be made with
the result of `tuneRanger`.

Out of curiosity I also fit `ranger` directly with the recommended
tuning from `tuneRanger` to compare their accuracy.

While preparing this, I notice `makeCLassifTask` has a `positive`
parameter to indicate the positive class, so I put this in a repeatable
function and collect the metrics for both classes.

```{r}
compare_tuneRanger_and_ranger <- function(positive_class) {
    # 1. Create classification task
    task <- mlr::makeClassifTask(
        data     = train_df,
        target   = 'disease_present',
        positive = positive_class 
    )
    
    # 2. Get its estimated execution time (suppressing other output)
    sink(nullfile())
    etime <- estimateTimeTuneRanger(task)
    sink()
    
    # 3. Execute task, capturing time
    rtime <- system.time(
        res <- tuneRanger(
            task,
            measure        = list(acc), # or multiclass.brier,, auc, logloss, ...
            num.trees      = 1000,
            num.threads    = 2,
            iters          = 20,
            save.file.path = NULL,
            show.info      = FALSE
        )
    )

    # 4. Predict with tuneRanger model
    pred      <- predict(res$model, newdata = test_df)
    
    # 5. Calculate its accuracy
    tuned_acc <- pred$data |>
        as.data.frame() |>
        dplyr::summarize(acc = 100*sum(truth == response)/n()) |>
        pull(acc)
    
    # 6. Fit ranger with the recommended parameter settings
    fit <- ranger(
        fc,
        train_df,
        mtry            = res$recommended.pars$mtry,
        min.node.size   = res$recommended.pars$min.node.size,
        sample.fraction = res$recommended.pars$sample.fraction
    )

    # 7. Report it
    tribble(
        ~PositiveClass, ~Estimated, ~Actual,
        ~Estimated, ~tuneRanger, ~ranger,
        ~mtry, ~min.node.size, ~sample.fraction,
        positive_class, etime, rtime['elapsed'],
        100*res$recommended.pars$acc,
        tuned_acc,
        accuracy(fit, test_df),
        res$recommended.pars$mtry,
        res$recommended.pars$min.node.size,
        round(res$recommended.pars$sample.fraction, 4)
    )
}

# 8. Blend the results
class1 <- compare_tuneRanger_and_ranger(positive_class = 1)
class0 <- compare_tuneRanger_and_ranger(positive_class = 0)
rbind(class1, class0) |>
    kable(digits = 3) |>
    add_header_above(
        c(' ' = 1, 'Time' = 2, 'Accuracy' = 3, 'Recommendations' = 3),
        underline = TRUE
    ) |> 
    kable_styling(
        bootstrap_options = 'bordered',
        full_width = FALSE,
        position = 'left'
    )
```

The differences are negligible and vary between runs. I conclude there
is no effective difference.

## Positive Class, Factor Order

Here I make one more comparison - fitting and predicting with `ranger`
against `train_df` and `test_df` as-is, with implicit factor ordering of
0, 1; and then again with explicit factor-ordering of 1,0.

```{r}
which_class <- function(.train = train_df, .test = test_df) {
    fit  <- ranger(fc, .train)
    pred <- predict(fit, .test)
    cm   <- confusionMatrix(pred$predictions, factor(.test$disease_present))
    c(
        accuracy = 100*cm$overall[['Accuracy']],
        correct  = diag(cm$table) |> sort()
    )
}

zero_first <- which_class()
one_first  <- which_class(
    train_df |> mutate(disease_present = factor(disease_present, ordered=T, levels=c(1,0))),
    test_df  |> mutate(disease_present = factor(disease_present, ordered=T, levels=c(1,0)))
)
    
cbind(zero_first, one_first) |>
    kable(caption = 'Factor Order Effect?') |>
    kable_styling(full_width = FALSE, position = 'left')
```

Additional evidence for the same conclusion - differences are minimal
and variable.

## RandomForest Decision Tree

I like to get a glimpse into the "thinking" of these models, and for
that I look to a `randomForest` decision tree. Unfortunately there is a
trade-off between depth of information and clarity. To help, I shorten
the column names and hackishly round digits to two decimals.

```{r message = FALSE, fig.width = 14, fig.height = 14}
fit <- randomForest(
    factor(disease_present) ~ .,
    train_df |> dplyr::rename(h2o = rainfall, tmp = temperature, pH = soil_ph, hum = humidity),
    ntree = 20
)
simplest_tree <- which.min(fit$forest$ndbigtree)
fit$forest$xbestsplit <- round(fit$forest$xbestsplit, 2)  # 'digits' workaround
plot.getTree(
    fit,
    k     = simplest_tree,
    depth = 6,
    main  = 'RandomForest Decision Tree'
)
```

## RandomForest - Other Metrics

Here I look at metrics in addition to accuracy.

Prevalence was known from the [class distribution](#class_distribution),
the 75% portion of the positive class (0, healthy).

Sensitivity, or true positive, is high; and Specificity, or true
negative, is fairly low. It seems plausible that those numbers are
reflective of the 75/25% distribution.

I am uncertain what to conclude. Accuracy and precision seem pretty
good, but sensitivity and specificity are not well balanced.

```{r}
metrics <- c('Prevalence', 'Sensitivity', 'Specificity', 'Precision', 'F1')
fit   <- randomForest(factor(disease_present) ~ ., train_df, ntree = 20)
pred  <- predict(fit, test_df)
cm    <- confusionMatrix(pred, factor(test_df$disease_present))
cm$byClass |>
    as.data.frame() |>
    rownames_to_column() |>
    pivot_wider(names_from = rowname, values_from = `cm$byClass`) |>
    select(all_of(metrics)) |>
    mutate(
        Accuracy = cm$overall[['Accuracy']]*100,
        F_meas   = caret::F_meas(
            data      = factor(train_df$disease_present),
            reference = factor(test_df$disease_present)
        )
    ) |>
    knitr::kable(digits = 3, caption = 'Ranger Classification Metrics') |>
    kable_styling(bootstrap_options = 'bordered', full_width = FALSE, position = 'left')
```

# Conclusions

Once again, I have not been able to improve models significantly beyond
what I consider the baseline. It seems the models' default modes are
nearly ideal. But, as mentioned in the [Introduction](#introduction) I'm
gauging against [others on
Kaggle](https://www.kaggle.com/datasets/turakut/plant-disease-classification/code),
and my best of 87% rivals what I've seen.

My primary goal was to continue learning, and here in particular I
learned:

-   `sink` and `nullfile`, for something analogous to `> /dev/null`

    But - danger, danger - do not leave a lingering `sink` object!

-   `curl_download`, to pull down Kaggle data without API credentials

-   ellipsis (`...`) argument, used here to make my model framework
    smaller and more flexible

-   more `kable`, especially making narrow tables NOT full width, and
    left-aligned

-   `labeller`, for plotting meaningful words instead of 0,1

-   `tuneRanger` package and tools

-   `datatable`, which I used via Perl years ago

-   the effect, if any, of which class is positive

-   `tryCatch`, to make the framework more robust

This was my first major exposure to library conflicts. I believe the
only instance I had seen prior was needing to fully qualify
`dplyr::select`. Here, I also had to qualify `ggcorrplot::cor_pmat` and
`caret::train`, neither of which was the least bit obvious; and later
`dplyr::rename` and `dplyr::mutate`. I am beginning to think that
fully-qualifying is a good practice.

This project turned up a few new interesting topics that I will revisit
in upcoming projects.

-   Using [tidymodels](https://tune.tidymodels.org) and associated
    packages to create tuning recipes and workflows, which seems to be
    an alternative to `train`
-   Using
    [dtree](https://www.rdocumentation.org/packages/dtree/versions/0.4.2/topics/dtree),
    for what seems to provide a simple ensemble of decision-tree models
-   Perhaps [function
    factories](https://adv-r.hadley.nz/function-factories.html) to
    reduce near-redundant code in my framework

# References

<https://stackoverflow.com/questions/48764060/portable-null-device-in-r>

[https://medium.com/\@asimango/the-positive-class-what-should-it-be-in-a-machine-learning-binary-classification-problem-36c316da1127](https://medium.com/@asimango/the-positive-class-what-should-it-be-in-a-machine-learning-binary-classification-problem-36c316da1127){.uri}

[https://adv-r.hadley.nz/function-factories.html#function-factories](https://adv-r.hadley.nz/function-factories.html){.uri}

[https://medium.com/\@ravi.abhinav4/improving-class-imbalance-with-class-weights-in-machine-learning-af072fdd4aa4](https://medium.com/@ravi.abhinav4/improving-class-imbalance-with-class-weights-in-machine-learning-af072fdd4aa4){.uri}

<https://www.tidyverse.org/blog/2022/05/case-weights/>
