---
title: 'UCI Mushrooom Analysis'
author: hbmaclean
date: "`r Sys.Date()`"
editor_options:
  markdown:
    toc: true
    wrap: 72
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: show
---

```{r setup, include=FALSE}
repo <- "http://cran.us.r-project.org"
packages <- c(
    'caret', 'stringr', 'readr', 'dplyr', 'tidyr', 'tibble', 'hash',
    'kableExtra', 'grid', 'gridExtra', 'factoextra', 'GGally', 'ranger',
    'varhandle', 'devtools', 'randomForest', 'MASS'
)
for (package in packages) {
    if (!require(package, character.only = TRUE)) {
        install.packages(package, repos = repo)
        library(package, character.only = TRUE)
    }
}
```

# Introduction

For this Data Science capstone project, I have chosen to analyze
[mushroom data](https://archive.ics.uci.edu/dataset/73/mushroom) from
[UCI.edu](https://uci.edu) - to attempt to predict whether a mushroom is
poisonous or edible, given color, odor, cap, stalk, habitat, and other
descriptions.

> From Audobon\<sic\> Society Field Guide; mushrooms described in terms
> of physical characteristics; classification: poisonous or edible This
> data set includes descriptions of hypothetical samples corresponding
> to 23 species of gilled mushrooms in the Agaricus and Lepiota Family
> (pp. 500-525). Each species is identified as definitely edible,
> definitely poisonous, or of unknown edibility and not recommended.
> This latter class was combined with the poisonous one. The Guide
> clearly states that there is no simple rule for determining the
> edibility of a mushroom; no rule like "leaflets three, let it be" for
> Poisonous Oak and Ivy.

I have no expertise in the area, but know enough about mushrooms to be
intrigued. In addition to their culinary and medicinal uses, I know
mushrooms are critical for breaking down organic material; are used in
soil and water remediation projects like oil spills; and have numerous
symbiotic relationships, including with algae (together composing
[lichen](https://en.wikipedia.org/wiki/Lichen)) and [Indian
pipes](https://www.nybg.org/blogs/science-talk/2018/09/indian-pipes-the-parasitic-plants-in-bloom-at-the-new-york-botanical-garden/#:~:text=Indian%20pipes%20are%20unusual%20in,be%20inaccessible%20to%20the%20plants.)
(a wildflower that does not photosynthesize.)

I did harvest and eat mushrooms once - a large, textbook cluster of
[Grifola frondosa](https://en.wikipedia.org/wiki/Grifola_frondosa),
Hen-of-the-woods, growing at the base of an oak tree in my yard. Once
dried, it added a nice earthy flavor to soups and other dishes.

\newpage

# Data Creation

This downloads the
[zip](https://archive.ics.uci.edu/dataset/73/mushroom), extracts the
data file, reads and parses the lines into a `data.frame` and adds
column names.

```{r}
# Download mushroom.zip from UCI.edu and render it as a data.frame
fetch_data <- function(verbose=FALSE) {
    options(timeout = 120)
    tmp   <- tempfile()
    zip   <- "https://archive.ics.uci.edu/static/public/73/mushroom.zip"

    download.file(zip, tmp, quiet = !verbose)
    data <- as.data.frame(
        str_split(
            read_lines(
                unz(tmp, "agaricus-lepiota.data")
            ),
            ",",
            simplify=TRUE
        )
    )

    colnames(data) <- c(
        "poisonous", "cap_shape", "cap_surface", "cap_color",
        "bruises", "odor", "gill_attachment", "gill_spacing",
        "gill_size", "gill_color", "stalk_shape", "stalk_root",
        "stalk_surface_above_ring", "stalk_surface_below_ring",
        "stalk_color_above_ring", "stalk_color_below_ring",
        "veil_type", "veil_color", "ring_number", "ring_type",
        "spore_print_color", "population", "habitat"
    )
    return(data)
}

mushroom_df <- fetch_data()
mushroom_df |> head() |> dplyr::select(poisonous, cap_shape, gill_color, odor, habitat)
```

\newpage

# Data Pre-Processing {#data_preprocessing}

For much analysis, numeric data is supportive of more tools. But for
plotting, I'll want human-friendly data.

Making numeric is a simple matter of matching `letters`, with special
attention toward the the first column, the outcome.

```{r}
# Convert single-letter abbreviations to numbers for all but first column.
# For the first column, convert 'p' (poisonous) to 1, otherwise 0.
# Some values are '?', which obviously do not map to letters and are coalesced to 0
make_numeric <- function(data) {
    data |>
        rowwise() |>
        mutate(across(-c(1), ~ coalesce(match(., letters), 0))) |>
        mutate(poisonous = ifelse(poisonous == 'p', 1, 0)) |>
        ungroup()
}

numeric_df <- make_numeric(mushroom_df)
numeric_df |> head() |> dplyr::select(poisonous, cap_shape, gill_color, odor, habitat)
```

\newpage

To make the data human-friendly, there is a dictionary of terms provided
on the [website](https://archive.ics.uci.edu/dataset/73/mushroom). I
convert that to a `hash` lookup table:

```{r}
# Map letters to their definitions, provided on website
make_legible <- function(data) {
    library(hash)

    # colors and surfaces are used several times
    colors <- hash(
        keys   = c('k', 'n', 'b', 'h', 'g', 'r', 'o', 'p', 'u', 'e', 'w', 'y', 'c'),
        values = c('black', 'brown', 'buff', 'chocolate', 'gray', 'green', 'orange',
                   'pink', 'purple', 'red', 'white', 'yellow', 'cinnamon')
    )

    surfaces <- hash(
        keys   = c('f', 'y', 'k', 's', 'g'),
        values = c('fibrous', 'scaly', 'silky', 'smooth', 'grooves')
    )

    values_map  <- hash(
        keys    = names(data),
        values = c(
            hash(
               keys   = c('p', 'e'),
               values = c(1, 0)
            ),
            hash(
               keys   = c('b', 'c', 'x', 'f', 'k', 's'),
               values = c('bell', 'conical', 'convex', 'flat', 'knobbed', 'sunken')
            ),
            surfaces,
            colors,
            hash(
               keys   = c('t', 'f'),
               values = c('bruises', 'no')
            ),
            hash(
               keys   = c('a', 'l', 'c', 'y', 'f', 'm', 'n', 'p', 's'),
               values = c('almond', 'anise', 'creosote', 'fishy', 'foul',
                          'musty', 'none', 'pungent', 'spicy')
            ),
            hash(
               keys   = c('a', 'd', 'f', 'n'),
               values = c('attached', 'descending', 'free', 'notched')
            ),
            hash(
               keys   = c('c', 'w', 'd'),
               values = c('close', 'crowded', 'distant')
            ),
            hash(
               keys   = c('b', 'n'),
               values = c('broad', 'narrow')
            ),
            colors,
            hash(
               keys   = c('e', 't'),
               values = c('enlarging', 'tapering')
            ),
            hash(
               keys   = c('b', 'c', 'u', 'e', 'z', 'r', '?'),
               values = c('bulbous', 'club', 'cup', 'equal', 'rhizomorphs',
                          'rooted', 'missing')
            ),
            surfaces,
            surfaces,
            colors,
            colors,
            hash(
               keys   = c('p', 'u'),
               values = c('partial', 'universal')
            ),
            colors,
            hash(
               keys   = c('n', 'o', 't'),
               values = c('none', 'one', 'two')
            ),
            hash(
               keys   = c('c', 'e', 'f', 'l', 'n', 'p', 's', 'z'),
               values = c('cobwebby', 'evanescent', 'flaring', 'large',
                          'none', 'pendant', 'sheathing', 'zone')
            ),
            colors,
            hash(
               keys   = c('a', 'c', 'n', 's', 'v', 'y'),
               values = c('abundant', 'clustered', 'numerous',
                          'scattered', 'several', 'solitary')
            ),
            hash(
               keys   = c('g', 'l', 'm', 'p', 'u', 'w', 'd'),
               values = c('grasses', 'leaves', 'meadows', 'paths',
                          'urban', 'waste', 'woods')
            )
        )
    )
    data |>
        rowwise() |>
        mutate(across(everything(), ~ values_map[[cur_column()]][[.]])) |>
        ungroup()
}
```

\newpage

```{r}
verbose_df <- make_legible(mushroom_df)
verbose_df |> head() |> dplyr::select(poisonous, cap_shape, gill_color, odor, habitat)
```

\newpage

# Data Analysis

This section is divided into [Packaged Tools
Analysis](#packaged_tool_analysis), using available tools such as
`prcomp`, `nearZeroVariance`, `aov`, and more; and [Custom
Analysis](#custom_analysis), a `dplyr` look into data trends.

## Packaged Tools Analysis {#packaged_tool_analysis}

Here I use readily available tools to primarily determine which
predictors are most influential and which can be omitted from analysis.
Admittedly, I don't have a firm enough understanding of many of these
tools to make confident conclusions; and am learning through usage.

### Principal Component Analysis {#principal_component_analysis}

First, I look for PCs that can be ignored.

```{r}
pca <- prcomp(numeric_df)
which(cumsum(pca$sdev^2)/sum(pca$sdev^2) >= 0.90)
```

Evidently the first 10 PCs explain 90% of the overall variation.

\newpage

**PC Rotation Barchart**

Here, for clarity, I plot only the top 5 PCs - which, from
`summary(pca)` are known to comprise about 19%, 13%, 12%, 11%, and 10%
of the overall variance:

```{r}
barchart(pca$rotation[,1:5], auto.key = TRUE)
```

To interpret this, I found it useful to also look at the output of
`round(scale(pca$rotation[,1:5]),3)`; and I also read [this
conversation](https://stats.stackexchange.com/questions/144508/conclusions-from-output-of-a-principal-component-analysis).
The plot indicates that, for example, `gill_color`, `ring_type`,
`bruises`, and `cap_color` are strongly and directly related to PC1,
while `spore_print_color`, `gill_size`, `stalk_shape`, and `odor` are
inversely related. We can also see that `veil_type` and
`gill_attachment` have little sway, as they have only 1 and 2 distinct
values, respectively.

\newpage

**PC Variance Explained**

And here is the %variance for the top 10 PCs, hand-rolled and via
`factoextra::fviz_eig` - evidently known as a [Scree
plot](https://en.wikipedia.org/wiki/Scree_plot):

```{r}
s   <- summary(pca)
p1  <- s$importance[,1:10] |>
    as.data.frame() |>
    rownames_to_column() |>
    filter(rowname == 'Proportion of Variance') |>
    pivot_longer(-rowname) |>
    dplyr::select(-rowname) |>
    mutate(name = reorder(name, desc(value))) |>
    ggplot(aes(name, value)) +
    geom_col() +
    labs(x = '', y = 'Variance Explained') +
    theme(axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)) +
    scale_y_continuous(labels = scales::percent)

library(factoextra)
p2  <- fviz_eig(pca)

grid.arrange(p1, p2, ncol = 2)
```

It is evident that much of the variance is explained by the first five
PCs, with significance declining thereafter.

\newpage

**PC Importance**

Below are the column name, most/least correlating columns, and
%importance for each PC explaining at least 11% of the total variance
(11% chosen for PDF width):

```{r}
sapply(which(s$importance[2,] > .11), function(i) {
    c(
        column     = names(numeric_df)[i],
        max        = names(numeric_df)[which.max(pca$rotation[,i])],
        min        = names(numeric_df)[which.min(pca$rotation[,i])],
        importance = sprintf('%%%.2f', s$importance[2,i]*100)
    )
})
```

Again also eyeballing the output of
`round(scale(pca$rotation[,1:5]),3)`, the two are in agreement. For
example, for PC1, `gill_color` is largest (2.242) and
`spore_print_color` is smallest (-1.666).

\newpage

**PC Variance Contribution**

Here I plot the variance contribution of each column for PC1, via
`prcomp` results and code derived from [STHDA (Statistical Tools for
High-Throughput Data
Analysis)](https://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp):

```{r}
var_coord_func <- function(loadings, comp.sdev){
    loadings*comp.sdev
}

loadings    <- pca$rotation
sdev        <- pca$sdev
var.coord   <- t(apply(loadings, 1, var_coord_func, sdev))
var.cos2    <- var.coord^2
comp.cos2   <- apply(var.cos2, 2, sum)
contrib     <- function(var.cos2, comp.cos2){var.cos2*100/comp.cos2}
var.contrib <- t(apply(var.cos2,1, contrib, comp.cos2))

var.contrib[,1] |>
    as.data.frame() |>
    rownames_to_column(var = 'column') |>
    mutate(value  = scale(pick(2)), column = reorder(column, value)) |>
    dplyr::select(column, value) |>
    ggplot(aes(value, column)) + labs(x = '', y = '') +
    geom_col()
```

This does seem to illustrate the composition of PC1 clearly.

\newpage

### Near-Zero Variance {#near_zero_variance}

This indicates three columns have near-zero variance. (Assistance
[here](https://stackoverflow.com/questions/6347356/creating-a-comma-separated-vector)
and
[here](https://stackoverflow.com/questions/28757372/printing-line-breaks-using-sprintf-with-shiny)
for `cat` and `shQuote`.)

```{r}
indices <- nearZeroVar(numeric_df)
columns <- names(numeric_df)[c(indices)]
cat(sprintf(
    "Indices: %s\nColumns: %s",
    paste(shQuote(indices), collapse=", "),
    paste(shQuote(columns), collapse=", ")
))
```

The latter two columns have already been deemed unnecessary via PCA.
Column 7, `gill_attachment`, has the following distribution:

```{r}
table(numeric_df[,7])
```

From the `value_map` in [Data Pre-Processing](#data_preprocessing)
coupled with mapping letters to numbers, we can derive that "1" equals
"a" (attached) and "6" equals "f" (free) - so 2.5% of gills are attached
and 97.5% are free.

\newpage

### Variance, Covariance, Correlation {#correlation_and_covariance}

Here, I try boxplots of `var` and `cov`. I did `cor` too, but it looks
identical to `var` so I've omitted it. I compute them first; but plot
them separately as they do not arrange well with `grid.arrange`. (I also
read [this
article](https://www.countbayesie.com/blog/2015/2/21/variance-co-variance-and-correlation)
as a refresher on variance, covariance and correlation.)

```{r warning=FALSE}
self_correlation <- 1
common_layers <- list(
    theme(legend.position = ''),
    geom_boxplot(),
    geom_hline(aes(yintercept = 0, color = 'pink')),
    coord_flip()
)

variance <- cor(numeric_df) |>
    as.data.frame() |>
    rownames_to_column() |>
    dplyr::select(-rowname) |>
    pivot_longer(everything()) |>
    filter(value != self_correlation) |>
    group_by(name) |> mutate(avg = mean(value)) |> ungroup() |>
    mutate(name = reorder(name, desc(avg))) |>
    ggplot(aes(name, value)) +
    labs(x = '', y = '', title = 'Variance') +
    common_layers

covariance <- cov(numeric_df) |>
    as.data.frame() |>
    rownames_to_column() |>
    dplyr::select(-rowname) |>
    pivot_longer(everything()) |>
    filter(value != self_correlation) |>
    group_by(name) |> mutate(avg = mean(value)) |> ungroup() |>
    mutate(name = reorder(name, desc(avg))) |>
    ggplot(aes(name, value)) +
    labs(x = '', y = '', title = 'Covariance') +
    common_layers
```

\newpage

```{r echo = FALSE}
variance
```

From this, I would guess that `poisonous` has primarily *negative*
correlations with its predictors, with its mean below zero; and that
`stalk_color_*` have most definitive correlations, with their
inter-quartile ranges completely above zero. The outliers are about
evenly split above- and below zero.

```{r echo = FALSE}
covariance
```

Note that `veil_type` was NA in Variance but joins the mix here, as a
covariant. The outlier ratio shifted - many more are now above zero. And
in addition to `stalk_color_*`, `stalk_surface_*` and `cap_surface` are
also almost wholly above zero.

\newpage

### ANOVA (aov) {#aov}

`aov` suggests that `bruises`, `gill_spacing`, `gill_size` are most
influential.

```{r}
unclass(summary(stats::aov(poisonous ~ ., data = numeric_df[c(1:6,8:10)]))) |>
    as.data.frame() |>
    rownames_to_column(var = 'Column') |>
    filter(!is.na(F.value)) |>
    mutate(Column = reorder(Column, desc(F.value))) |>
    ggplot(aes(Mean.Sq, F.value, color = Column)) +
    geom_point(cex=4) +
    scale_x_log10()
```

\newpage

### GGPAIRS {#ggpairs}

Here I plot the correlations for what seem to be the most influential
predictors. (The plot becomes unwieldy with too many predictors.) I'm
still learning about this tool, not ready to make any conclusions from
it yet.

```{r message=FALSE}
ggpairs(numeric_df[, c(1:3,5,8:10)])
```

### Packaged Tools Conclusion

From a variety of packaged tools, I conclude that three columns can be
omitted due to [near-zero variation](#near_zero_variance) and [13
PCs](#principal_component_analysis) can be omitted as the other 10
explain 90% of the variation. If I end up using `train`, at least `pca`
and `nzv` should be worthwhile pre-options, with experimental tuning. I
also have some sense of the more influential columns, which can be a
starting point for subsequent analysis.

\newpage

## Custom Analysis {#custom_analysis}

Here I inspect the data with SQL-like `dplyr` methods, looking for
telltale patterns.

First, i define three layers that will be used verbatim throughout the
plots:

```{r}
common_layers <- list(
     geom_bar(stat = 'identity'),
     scale_y_continuous(labels = scales::percent),
     theme(
         legend.position = '',
         axis.text.x = element_text(angle = 60, vjust = 1, hjust = 1)
     )
)
```

### Mushroom Cap Attributes

Here I make bar plots of the poisonous percentage for each column value
of the cap-related columns.

```{r}
cap_shape <- aggregate(verbose_df, poisonous ~ cap_shape, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(cap_shape, poisonous, fill = cap_shape)) +
      labs(x = "", y = "", title = "Shape") +
      common_layers

cap_surface <- aggregate(verbose_df, poisonous ~ cap_surface, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(cap_surface, poisonous, fill = cap_surface)) +
      labs(x = "", y = "", title = "Surface") +
      common_layers

cap_color <- aggregate(verbose_df, poisonous ~ cap_color, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(cap_color, poisonous, fill = cap_color)) +
      labs(x = "", y = "", title = "Color") +
      common_layers

grid.arrange(
    cap_shape, cap_surface, cap_color,
    ncol = 3,
    top = textGrob(
        "Poison Percentage by Cap Attributes",
        gp = gpar(fontsize = 12, font = 3)
    )
)
```

\newpage

### Mushroom Gill Attributes

Here, I do the same for gill-related columns. Note, `gill_attachment`
has been shown to have [near-zero variance](#near_zero_variance), and
will likely be omitted in final analysis.

```{r}
gill_attachment <- aggregate(verbose_df, poisonous ~ gill_attachment, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(gill_attachment, poisonous, fill = gill_attachment)) +
      labs(x = "", y = "", title = "Attachment") +
      common_layers

gill_spacing <- aggregate(verbose_df, poisonous ~ gill_spacing, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(gill_spacing, poisonous, fill = gill_spacing)) +
      labs(x = "", y = "", title = "Spacing") +
      common_layers

gill_size <- aggregate(verbose_df, poisonous ~ gill_size, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(gill_size, poisonous, fill = gill_size)) +
      labs(x = "", y = "", title = "Size") +
      common_layers

gill_color <- aggregate(verbose_df, poisonous ~ gill_color, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(gill_color, poisonous, fill = gill_color)) +
      labs(x = "", y = "", title = "Color") +
      common_layers

grid.arrange(
    gill_attachment, gill_spacing, gill_size, gill_color,
    ncol = 4,
    top = textGrob(
        "Poison Percentage by Gill Attributes",
        gp = gpar(fontsize = 12, font = 3)
    )
)
```

\newpage

### Mushroom Stalk Attributes

Here, I do the same for stalk-related columns.

```{r}
stalk_shape <- aggregate(verbose_df, poisonous ~ stalk_shape, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(stalk_shape, poisonous, fill = stalk_shape)) +
      labs(x = "", y = "", title = "Shape") +
      common_layers

stalk_root <- aggregate(verbose_df, poisonous ~ stalk_root, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(stalk_root, poisonous, fill = stalk_root)) +
      labs(x = "", y = "", title = "Root") +
      common_layers

stalk_surface_above <- aggregate(verbose_df, poisonous ~ stalk_surface_above_ring, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(stalk_surface_above_ring, poisonous, fill = stalk_surface_above_ring)) +
      labs(x = "", y = "", title = "Surface", subtitle = "Above Ring") +
      common_layers

stalk_surface_below <- aggregate(verbose_df, poisonous ~ stalk_surface_below_ring, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(stalk_surface_below_ring, poisonous, fill = stalk_surface_below_ring)) +
      labs(x = "", y = "", title = "Surface", subtitle = "Below Ring") +
      common_layers

stalk_color_above <- aggregate(verbose_df, poisonous ~ stalk_color_above_ring, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(stalk_color_above_ring, poisonous, fill = stalk_color_above_ring)) +
      labs(x = "", y = "", title = "Color", subtitle = "Above Ring") +
      common_layers

stalk_color_below <- aggregate(verbose_df, poisonous ~ stalk_color_below_ring, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(stalk_color_below_ring, poisonous, fill = stalk_color_below_ring)) +
      labs(x = "", y = "", title = "Color", subtitle = "Below Ring") +
      common_layers

grid.arrange(
    stalk_shape, stalk_root, stalk_surface_above,
    stalk_surface_below, stalk_color_above, stalk_color_below,
    ncol = 3,
    nrow = 2,
    top = textGrob(
        "Poison Percentage by Stalk Attributes",
        gp = gpar(fontsize = 12, font = 3)
    )
)
```

\newpage

### Mushroom Veil and Ring Attributes

Here I do the same for veil- and ring-related columns. Note, `veil_type`
and `veil_color` have been shown to have [near-zero
variance](#near_zero_variance) and will likely be omitted from final
analysis.

```{r}
veil_type <- aggregate(verbose_df, poisonous ~ veil_type, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(veil_type, poisonous, fill = veil_type)) +
      labs(x = "", y = "", title = "Veil Type") +
      common_layers

veil_color <- aggregate(verbose_df, poisonous ~ veil_color, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(veil_color, poisonous, fill = veil_color)) +
      labs(x = "", y = "", title = "Veil Color") +
      common_layers

ring_type <- aggregate(verbose_df, poisonous ~ ring_type, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(ring_type, poisonous, fill = ring_type)) +
      labs(x = "", y = "", title = "Ring Type") +
      common_layers

ring_number <- aggregate(verbose_df, poisonous ~ ring_number, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(ring_number, poisonous, fill = ring_number)) +
      labs(x = "", y = "", title = "Ring Number") +
      common_layers

grid.arrange(
    veil_type, veil_color, ring_type, ring_number,
    ncol = 4,
    top = textGrob(
        "Poison Percentage by Veil and Ring Attributes",
        gp = gpar(fontsize = 12, font = 3)
    )
)
```

\newpage

### Mushroom Other Attributes

Here I do the same for a variety of one-off columns.

```{r}
bruises <- aggregate(verbose_df, poisonous ~ bruises, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(bruises, poisonous, fill = bruises)) +
      labs(x = "", y = "", title = "Bruises") +
      common_layers

odor <- aggregate(verbose_df, poisonous ~ odor, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(odor, poisonous, fill = odor)) +
      labs(x = "", y = "", title = "Odor") +
      common_layers

population <- aggregate(verbose_df, poisonous ~ population, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(population, poisonous, fill = population)) +
      labs(x = "", y = "", title = "Population") +
      common_layers

habitat <- aggregate(verbose_df, poisonous ~ habitat, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(habitat, poisonous, fill = habitat)) +
      labs(x = "", y = "", title = "Habitat") +
      common_layers

spore_print_color <- aggregate(verbose_df, poisonous ~ spore_print_color, sum) |>
      mutate(poisonous = poisonous/nrow(verbose_df)) |>
      ggplot(aes(spore_print_color, poisonous, fill = spore_print_color)) +
      labs(x = "", y = "", title = "Spore Color") +
      common_layers

grid.arrange(
    bruises, odor, population, habitat, spore_print_color,
    ncol = 3,
    top = textGrob(
        "Poison Percentage by Other Attributes",
        gp = gpar(fontsize = 12, font = 3)
    )
)
```

\newpage

### Custom Analysis Conclusion

The above research provides some direction. Here I calculate poison
percentage for promising patterns.

```{r}
verbose_df |>
    filter(
        gill_color == 'buff'
        # odor %in% c('fishy', 'foul', 'spicy')
        # etc.
    ) |> summarize(n=n(), pct = sum(poisonous)/n)
```

And here are the patterns I tried, and the number of rows returned and
percentage of poisonous:

```{r echo=FALSE, message=FALSE, warning=FALSE}
data <- tibble(
    ATTRIBUTE_SET = c(
        'Cap - convex/flat; scaly/smooth/fibrous; brown/gray/red/yellow',
        'Gills - buff-colored (*)',
        'Stalk - silky/smooth; white/pink/brown/buff',
        'Ring - one; evanescent/large/pendant',
        'Root - bulbous/missing',
        'Bruises - no',
        'Odor - fishy/foul/spicy (*)',
        'Population - several/solitary/scattered',
        'Habitat - grasses/leaves/paths/woods',
        'Spore Color - white/chocolate'
    ),
    NUMBER = c(
        5690, 1728, 6648, 7440, 6256, 4748, 3312, 7000, 7272, 4020
    ),
    POISON_PCT = c(
        48.7, 100, 56.1, 51.2, 57.8, 69.3, 100, 55.2, 49.6, 84.5
    ),
)
knitr::kable(data, caption = "Poison Tendencies") |>
    kable_styling(latex_options = c('striped')) |>
    footnote(symbol = c("Note: Buff-colored gills and fishy/foul/spicy odor overlap"))
```

\newpage

# Results

## First Model

I pieced together some of the conditions noted in [Custom
Analysis](#custom_analysis) and filtered ***out*** matching rows; then
iteratively looked for patterns in the remaining poisonous rows and
added them to the filtering; etc. until arriving at this:

```{r}
verbose_df |>
    mutate(
        guess = ifelse((
            odor %in% c('fishy', 'foul', 'spicy', 'creosote', 'pungent')
          | ring_number == 'none'
          | cap_shape == 'conical'
          | spore_print_color == 'green'
        ), 1, 0)
    ) |>
    summarize(accuracy= sum(guess==poisonous)/n()) |>
    pull(accuracy)
```

That seems too easy and very accurate. But the
[webpage](https://archive.ics.uci.edu/dataset/73/mushroom) indicates
that I should be near 100% - and I did not spot any pattern in the
balance of poisonous rows that do not match this query.

## Second Model

During the first project, I spent a lot of time with `train` and various
algorithms, only to sidestep the work because the data was too large.
Now I'd like to revisit; and `ranger` was one I spent most time on.

```{r}
index <- createDataPartition(y = numeric_df$poisonous, times = 1, p = 0.5, list = FALSE)
train <- numeric_df[index,]
test  <- numeric_df[-index,]
fit   <- ranger(factor(poisonous) ~ ., train)
pred  <- predict(fit, test)
cm    <- confusionMatrix(pred$predictions, factor(test$poisonous))

sprintf('Accuracy: %%%.2f', cm$overall[['Accuracy']]*100)
```

\newpage

## Final Model

What?! No tuning?

All that's left to do is run it against the full set:

```{r}
fit   <- ranger(factor(poisonous) ~ ., numeric_df)
pred  <- predict(fit, numeric_df)
cm    <- confusionMatrix(pred$predictions, factor(numeric_df$poisonous))

sprintf('Accuracy: %%%.2f', cm$overall[['Accuracy']]*100)
```

# Conclusion

I envisioned laboriously fine-tuning `ranger` or other algorithm to
reach a decent score. This was actually disappointing; below is the
brief progression of models:

```{r echo = FALSE}
data <- tibble(
    MODEL    = c("first", "second", "final"),
    ACCURACY = c(0.9945839, 1, 1)
)
knitr::kable(data, caption = "Model Progression") |> kable_styling()
```

Overall, I did not learn much with this project other than additional R
skills - particularly:

-   `aggregate` and `rowwise`, coupled with `mutate` and `across`
-   adding titles to `grid.arrange` and styling to `kable`
-   `hash`, a fundamental feature in Perl

I also carried over two lessons from the first project - to NOT use Rmd
markup tables, which do not `knit` well, nor headings below 3, which do
not `knit` to my taste.

\newpage

## Decision Tree {#decision_tree}

Finally, it occurred to me that if I had solved this immediately with
out-of-the-box `ranger` without first taking the `dplyr` approach, I
would have no idea how the algorithm might make its determinations -
i.e., what its decision tree might resemble. With that in mind, I found
[this](https://stats.stackexchange.com/questions/41443/how-to-actually-plot-a-sample-tree-from-randomforestgettree)
about `randomForest` and tried it. Note, I set `depth` for legibility,
at the expense of not reaching all leaf nodes.

```{r message=FALSE}
install_github('munoztd0/reprtree')
eval(substitute(library(pkg), list(pkg='reprtree')))
library(reprtree)

fit <- randomForest(factor(poisonous) ~ ., verbose_df)
cm  <- confusionMatrix(pred$predictions, factor(numeric_df$poisonous))
k   <- which.min(fit$forest$ndbigtree) # simplest tree
plot.getTree(
    fit,
    k     = k,
    depth = 6,
    main  = sprintf(
        '%s (%%%.2f accuracy)',
        'RandomForest Decision Tree',
        cm$overall[['Accuracy']]*100
    )
)
```

\newpage

# References

<https://dplyr.tidyverse.org/articles/rowwise.html>

[https://www.benjaminbell.co.uk/2018/03/principal-components-analysis-pca-in-r-part-2.html](http://127.0.0.1:46983/#0 "https://www.benjaminbell.co.uk/2018/03/principal-components-analysis-pca-in-r-part-2.html")

[https://www.modernstatisticswithr.com/regression.html](http://127.0.0.1:46983/#0 "https://www.modernstatisticswithr.com/regression.html")

<https://stackoverflow.com/questions/14726078/changing-title-in-multiplot-ggplot2-using-grid-arrange>

[https://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp](http://127.0.0.1:45647/#0 "https://www.sthda.com/english/articles/31-principal-component-methods-in-r-practical-guide/118-principal-component-analysis-in-r-prcomp-vs-princomp")

<https://haozhu233.github.io/kableExtra/awesome_table_in_pdf.pdf>

<https://stats.stackexchange.com/questions/41443/how-to-actually-plot-a-sample-tree-from-randomforestgettree>

<https://stats.stackexchange.com/questions/144508/conclusions-from-output-of-a-principal-component-analysis>

<https://stackoverflow.com/questions/18747091/reuse-ggplot-layers-in-multiple-plots/44721060>

<https://www.countbayesie.com/blog/2015/2/21/variance-co-variance-and-correlation>

<https://stackoverflow.com/questions/6347356/creating-a-comma-separated-vector>

<https://stackoverflow.com/questions/28757372/printing-line-breaks-using-sprintf-with-shiny>

\newpage

# Appendix

While the movielens project was graded by peers, this project was graded
by staff. Below is their feedback and my subsequent work to address it.

## Staff Feedback

Staff response was constructive, positive with suggestions (to which I
added numbers).

> This data science capstone project demonstrates strong exploratory
> data analysis and visualization work. You demonstrate impressive R
> programming skills, effectively transforming categorical mushroom
> data, and creating comprehensive and clear visualizations across all
> mushroom attributes. Good understanding of various R packages and
> statistical techniques, progressing logically from simple pattern
> recognition to more sophisticated modeling approaches using
> randomForest algorithms.
>
> There are some areas for enhancement. Noteably, the project (1) lacks
> acknowledgment of class imbalance despite achieving 100% accuracy,
> which is a strong indicator of imbalanced data that should be
> explicitly examined and reported. You want to provide class
> distribution counts and (2) consider metrics beyond accuracy
> (precision, recall, F1-score). (3) Data scaling was not applied before
> PCA analysis, which is essential for distance-based methods. (4) You
> want to establish simpler baselines like logistic regression, which
> would be appropriate for this binary classification problem. (5) The
> project lacks specific limitations discussion and future work
> recommendations.

## Response

1.  *Acknowledgement of class imbalance.*

    The many bar plots indicate a near-even balance of the two classes,
    but I should have been explicit.

    ```{r}
    # prop.table(table(numeric_df$poisonous)) # simpler, but less PDF-friendly
    numeric_df |>
        summarize(Poisonous = sum(poisonous)/n(), Harmless = 1-Poisonous) |>
        knitr::kable(caption = 'Poisonous Distribution') |>
        kable_styling()
    ```

    \newpage

2.  *Consider metrics beyond accuracy.*

    I thought 100% accuracy was enough if the classes are reasonably
    balanced. Here are other metrics:

    ```{r}
    metrics <- c('Sensitivity', 'Specificity', 'Prevalence', 'Recall', 'Precision')
    fit     <- ranger(factor(poisonous) ~ ., numeric_df)
    pred    <- predict(fit, numeric_df)
    cm      <- confusionMatrix(pred$predictions, factor(numeric_df$poisonous))
    round(c(
        cm$byClass[metrics],
        F1 = caret::F_meas(
            data      = factor(pred$predictions),
            reference = factor(numeric_df$poisonous)
        )
    ),3) |>
        as.data.frame() |>
        rownames_to_column() |>
        pivot_wider(names_from = rowname, values_from = c(2)) |>
        knitr::kable(caption = 'Ranger Classification Metrics') |> kable_styling()
    ```

3.  *Data scaling for PCA.*

    Fair point; though here it seems to have a punitive effect, with
    more PCs needed to explain 90% variance? To scale, `veil_type` must
    be omitted or `prcomp` complains, "cannot rescale a constant/zero
    column to unit variance":

    ```{r}
    columns  <- c(1:16,18:23)
    pca      <- prcomp(numeric_df[,columns], scale. = FALSE)
    unscaled <- which(cumsum(pca$sdev^2)/sum(pca$sdev^2) < 0.90)
    pca      <- prcomp(numeric_df[,columns], scale. = TRUE)
    scaled   <- which(cumsum(pca$sdev^2)/sum(pca$sdev^2) < 0.90)

    knitr::kable(
        tibble(
            SCALING = c('unscaled', 'scaled'),
            'PCs Explaining 90% Variance' = c(
                paste(unscaled, collapse = ', '),
                paste(scaled, collapse = ', ')
            )
        ),
        caption = 'Data Scaling for PCA'
    ) |> kable_styling()
    ```

    \newpage

4.  *Simpler baselines like logistic regression.*

    Simpler than querying with `dplyr`? And wait, regression? Surely
    this is a classification case, with two discrete outcomes? OK...:

    ```{r message = FALSE, warning = FALSE}
    fit     <- glm(poisonous ~ ., data = numeric_df, family = 'binomial')
    pred    <- predict(fit, numeric_df)
    y_hat   <- factor(ifelse(pred < .5, 0, 1))
    cm      <- confusionMatrix(y_hat, factor(numeric_df$poisonous))

    round(c(
        cm$byClass[metrics],
        F1 = caret::F_meas(
            data      = y_hat,
            reference = factor(numeric_df$poisonous)
        )
    ),3) |>
        as.data.frame() |>
        rownames_to_column() |>
        pivot_wider(names_from = rowname, values_from = c(2)) |>
        knitr::kable(caption = 'Logistic Regression with GLM') |> kable_styling()
    ```

5.  *Lacks discussion of limitations and future work considerations.*

    Given 100% accuracy plus all the other affirmative metrics of the
    final solution, I do not know what to say other than that I see no
    limitations or future work considerations.
